# 16S Dataset

## 📦 Build the phyloseq object (16S)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/16S"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Load metadata
metadat <- readxl::read_xlsx(file.path(datadir, "metadata_16S.xlsx")) %>%
  mutate(
    Date = as.Date(as.numeric(Date), origin = "1899-12-30"),
    Time = as.POSIXct(as.numeric(Time) * 86400, origin = "1970-01-01", tz = "UTC"),
    Time = format(Time, "%H:%M:%S"),
    Water_mass = as.factor(Water_mass)
  ) %>% as.data.frame()

rownames(metadat) <- rownames(seqtab.nochim)

# Create a Phyloseq object

ps_16s <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))
  
cat("🔹 A summary of the phyloseq object \n")
ps_16s

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_16s))
names(dna) <- taxa_names(ps_16s)
ps_16s <- merge_phyloseq(ps_16s, dna)
taxa_names(ps_16s) <- paste0("ASV", seq(ntaxa(ps_16s)))

```
```{r, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 These are the metadata variables:\n")
print(sample_variables(ps_16s))
```

### Positive and negative control

:::: {.columns}

::: {.column width="40%"}
```{r, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

ps_controls <- subset_samples(
  ps_16s,
  Name %in% c("65_zymo_positive", "66_Extraction_negative")
)

ps_controls <- prune_taxa(taxa_sums(ps_controls) > 0, ps_controls)

kable(sample_sums(ps_controls), caption = "Number of reads in the controls") %>% kable_styling(full_width = FALSE, position = "center")
```
:::

::: {.column width="20%"}
:::

::: {.column width="40%"}

```{r, message=FALSE, warning=FALSE}

kable(estimate_richness(ps_controls, measures = "Observed"), caption = "Observed richness") %>% kable_styling(full_width = FALSE, position = "center")

```
:::

::::

::: {.callout-note}
### 💬 Discussion — Controls

Excellent! Those control numbers are actually very reassuring:

**Positive control**: 733,263 reads, 14 ASVs → perfectly reasonable (mock community). However, it's a bit more than expected

**Extraction negative**: 4 reads, 4 ASVs → essentially clean (background noise).
  
:::

```{r, message=FALSE, warning=FALSE}

df_controls <- psmelt(ps_controls)

df_controls %>%
  group_by(Sample, Family) %>%
  summarise(Abundance = sum(Abundance)) %>%
  arrange(Sample, desc(Abundance)) %>%
  head(10) %>% kable() %>% kable_styling(full_width = FALSE)

```

::: {.callout-note}
### 💬 Discussion — Positive Controls : **ZymoBIOMICS Microbial Standard**

From the table above, we can ignore the ASVs that have low abundance (noise). We therefore have ASVs that match 7 families.

From the Zymo standard, we are expecting 8 bacteria:

🦠 Bacteria species:
- Pseudomonas aeruginosa
- Escherichia coli
- Salmonella enterica
- Lactobacillus fermentum
- Enterococcus faecalis
- Staphylococcus aureus
- Listeria monocytogenes
- Bacillus subtilis

:::

```{r, message=FALSE, warning=FALSE}
df_controls %>%
  group_by(Sample, Genus) %>%
  summarise(Abundance = sum(Abundance)) %>%
  arrange(Sample, desc(Abundance)) %>%
  head(9) %>% kable() %>% kable_styling(full_width = FALSE)

```
::: {.callout-note}
### 💬 Discussion — Controls Conclusion

The ZymoBIOMICS Microbial Community Standard positive control performed as expected. All eight bacterial genera included in the mock community were successfully recovered.The relative abundances of recovered taxa were broadly consistent with expected composition, although minor deviations were observed, likely reflecting known amplification biases associated with 16S rRNA gene primers.

A small proportion (~1%) of reads were assigned to Enterobacteriaceae without genus-level resolution, likely reflecting limitations of the amplified 16S region or conservative classification thresholds.

Overall, these results confirm the reliability of amplification, sequencing, and taxonomic assignment in the 16S workflow.

We can now remove the controls from the phyloseq object and move on with the ecological interpretation.
:::

```{r, message=FALSE, warning=FALSE}
ps_16s <- subset_samples(
  ps_16s,
  !Name %in% c("65_zymo_positive", "66_Extraction_negative")
)

# Prune zero-abundance taxa
ps_16s <- prune_taxa(taxa_sums(ps_16s) > 0, ps_16s)

saveRDS(ps_16s, "data/ps_16s.rds")

```

### Removal of cholorplast and mitochondria reads

In 16S rRNA metabarcoding datasets, sequences originating from **chloroplasts** and **mitochondria** are commonly co-amplified due to the evolutionary origin of these organelles from bacteria. In marine surface waters, such as Greenland fjords, chloroplast-derived 16S sequences from phytoplankton can represent a substantial proportion of total reads, particularly when using universal primers targeting the V4 region (e.g. 515F–806R). While these sequences are biologically meaningful, they do not represent free-living prokaryotic communities, which are typically the focus of bacterial diversity analyses. If retained, chloroplast and mitochondrial reads can dominate sequencing libraries and obscure patterns in heterotrophic and autotrophic bacterial taxa, thereby reducing the resolution and interpretability of microbial community structure.

For this reason, we will remove reads classified as chloroplast or mitochondrial prior to downstream analyses, ensuring that the resulting dataset more accurately reflects the composition and diversity of the target prokaryotic community.

With dada2::assignTaxonomy() using SILVA v138.2, chloroplast and mitochondrial assignments are often found in these places:

 - Order == "Chloroplast"
 - Family == "Mitochondria"

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

tax_tab <- as.data.frame(tax_table(ps_16s))
otu_tab <- as(otu_table(ps_16s), "matrix")

# make sure taxa are rows
if (!taxa_are_rows(ps_16s)) {
  otu_tab <- t(otu_tab)
}

# identify chloroplast and mitochondria ASVs across all taxonomic ranks
is_chloro <- apply(tax_tab, 1, function(x) any(grepl("Chloroplast", x, ignore.case = TRUE)))
is_mito   <- apply(tax_tab, 1, function(x) any(grepl("Mitochondria", x, ignore.case = TRUE)))

# total reads
total_reads <- sum(otu_tab)

# total reads assigned to each group
chloro_reads <- sum(otu_tab[is_chloro, , drop = FALSE])
mito_reads   <- sum(otu_tab[is_mito,   , drop = FALSE])

# percentages
chloro_pct <- 100 * chloro_reads / total_reads
mito_pct   <- 100 * mito_reads   / total_reads

cat("Total reads:", total_reads, "\n")
cat("Chloroplast reads:", chloro_reads, sprintf("(%.2f%%)\n", chloro_pct))
cat("Mitochondria reads:", mito_reads, sprintf("(%.2f%%)\n", mito_pct))
```

Overall, Chloroplast reads represent 18.5% of reads which is non negligible! But how are these reads distributed in our samples? Let's look at the amount of chloroplast and mitochondria reads per samples.

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# per-sample read counts
chloro_per_sample <- colSums(otu_tab[is_chloro, , drop = FALSE])
mito_per_sample   <- colSums(otu_tab[is_mito,   , drop = FALSE])
total_per_sample  <- colSums(otu_tab)

summary_df <- data.frame(
  Sample = colnames(otu_tab),
  Total_reads = total_per_sample,
  Chloroplast_reads = chloro_per_sample,
  Chloroplast_pct = 100 * chloro_per_sample / total_per_sample,
  Mitochondria_reads = mito_per_sample,
  Mitochondria_pct = 100 * mito_per_sample / total_per_sample
)

rownames(summary_df) <- NULL

kable(summary_df,
      caption = "Chloroplast and mitochondrial read summary",
      digits = 2) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "400px")
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# reshape to long format
plot_df <- summary_df %>%
  select(Sample, Chloroplast_pct, Mitochondria_pct) %>%
  pivot_longer(cols = c(Chloroplast_pct, Mitochondria_pct),
               names_to = "Type",
               values_to = "Percentage")

ggplot(plot_df, aes(x =Sample, y = Percentage, fill=Type)) +
  geom_col() +
  labs(x = "Sample",
       y = "Percentage  (%)",
       fill = "Read type",
       title = "Chloroplast and mitochondrial read proportions across samples") +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  )
```

The plot shows that chloroplast reads make up a substantial proportion of the dataset in many samples, in some cases approaching or exceeding roughly half of total reads, whereas mitochondrial reads remain consistently low across all samples. The marked variability among samples also suggests differences in phytoplankton biomass or composition across the samples. Chloroplasts are abundant mainly in surface and 15m samples.

These organelle-derived reads are ecologically informative because they reflect the presence of planktonic eukaryotes, particularly phytoplankton, within the sampled communities. However, because this part of the analysis is focused on the prokaryotic community, we will remove the chloroplast and mitochondrial sequences. These reads may nevertheless prove useful later when we will be investigating relationships between planktonic eukaryotes and prokaryotic community structure.

```{r, message=FALSE, warning=FALSE}
ps_16s <- subset_taxa(
  ps_16s,
  !Kingdom %in% c("Eukaryota", "NA") &
  !Order %in% c("Chloroplast") &
  !Family %in% c("Mitochondria")
)
```


## 📊 Data visualisation

In this section, we explore the community structure of the RoCSI 16S dataset using the
`phyloseq` object (`ps_16s`) generated above. We start by visualising read depth
per sample, followed by basic taxonomic composition summaries.

### Read depth per sample
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
read_depth <- data.frame(
Sample = sample_names(ps_16s),
Reads = sample_sums(ps_16s),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() + scale_y_continuous(labels = scales::label_number()) +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "Sequencing depth per sample",
x = "Sample",
y = "Number of non-chimeric reads"
)
```

### Rarefaction curve

Let's check whether sampling (sequencing) effort was sufficient to capture most of the observed ASV richness.

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
library(phyloseq)
library(vegan)

otu_tab <- as(otu_table(ps_16s), "matrix")

if (taxa_are_rows(ps_16s)) {
  otu_tab <- t(otu_tab)
}

rarecurve(otu_tab, step = 1000, cex = 0.5, label = FALSE)
```

The rarefaction curves show that most samples approach an asymptote as sequencing depth increases, indicating that the majority of detectable prokaryotic diversity has been captured.

The sample with **922 ASVs** (top curve) is sample **25_St9_D1**
The following down with **584 ASVs** (2nd top curve) is sample **03_St1_D45**

### Alpha diversity

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}


# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_16s)$Fjord <- as.factor(sample_data(ps_16s)$Fjord)

plot_richness(
  ps_16s,
  x = "Number",
  measures = c("Observed", "Shannon", "Simpson"),
  color = "Fjord", title = "16S Alpha diversity") +
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 4)]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns

Observed richness (number of ASVs), Shannon and Simpson diversity are fairly similar across fjords and the ocean samples.
  
:::

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}


# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_16s)$Fjord <- as.factor(sample_data(ps_16s)$Fjord)
sample_data(ps_16s)$Depth <- as.factor(sample_data(ps_16s)$Depth)

plot_richness(
  ps_16s,
  x = "Number",
  measures = c("Observed", "Shannon", "Simpson"),
  color = "Depth", title = "16S Alpha diversity") +
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 4)]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns

Observed richness (number of ASVs) tend to be increasing with depth.
:::

### Ordination

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
ord <- ordinate(ps_16s, method = "PCoA", distance = "bray")

# Create the ordination plot with ellipses
plot_ordination(ps_16s, ord, color = "Fjord") +
  geom_point(size = 3) +
  ggtitle("Ordination – 16S") +
  stat_ellipse(aes(group = Fjord), type = "norm", linetype = "dashed", size = 1) +
  theme_minimal()
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
ord <- ordinate(ps_16s, method = "PCoA", distance = "bray")

# Create the ordination plot with ellipses
plot_ordination(ps_16s, ord, color = "Depth") +
  geom_point(size = 3) +
  ggtitle("Ordination – 16S") +
  stat_ellipse(aes(group = Depth), type = "norm", linetype = "dashed", size = 1) +
  theme_minimal()
```

::: {.callout-note}
### 💬 Discussion — Beta diversity

The ordination shows **clustering of samples by fjord**, indicating fjord-level structuring of bacterial communities. Sermilik forms a relatively distinct and tightly clustered group, whereas Tunulliarfik exhibits greater within-fjord variability. Notably, there is considerable overlap between Itterdalk and Tunulliarfik, suggesting partial similarity in community composition and potentially shared environmental controls. Overall, **fjord identity contributes to community differentiation**, although separation is not absolute among all systems.

Next: check if dispersion differs significantly? 
betadisper(...)
permutest(...)

**NOTE**: some samples are pretty distant from the others . Who are they?
:::

::: {style="width:60%; margin:auto;"}
```{r, message=FALSE, warning=FALSE}
# Extract sample coordinates from the PCoA result
site <- as.data.frame(ord$vectors)
site$SampleID <- rownames(site)

# Get the 5 lowest values on Axis.1 (side of plot)
right5 <- site[order(site$Axis.1, decreasing = TRUE), ][1:5, c("SampleID", "Axis.1", "Axis.2")]
kable(right5) %>% kable_styling(full_width = FALSE, position = "center")
```
:::

::: {.callout-note}
The most divergent samples in the ordination correspond exclusively to surface samples (ID ending with D1). These stations were influenced by meltwater and low salinity conditions, suggesting that surface stratification and freshwater input may exert a strong effect on microbial community composition. This pattern indicates that vertical position and freshwater influence may be key drivers of community differentiation in these fjord systems.
:::


### Phylum composition

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",
               "#984EA3", "#A6D854", "#FFFF33", "#FF7F00", "#F781BF", "#999999",                  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F",
               "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")

library(microbiome)

# 1) Agglomerate to Phylum
ps_phylum <- tax_glom(ps_16s, taxrank = "Phylum")

# 2) Transform to relative abundance (compositional)
ps_rel <- microbiome::transform(ps_phylum, "compositional")

# 3) Top 20 taxa by mean relative abundance (returns taxa_names / OTU IDs)
top20 <- microbiome::top_taxa(ps_rel, n = 20)

# 4) Melt full object
df <- psmelt(ps_rel)

# Ensure Phylum label exists and is character (handles NAs)
df <- df %>%
  mutate(
    Phylum = as.character(Phylum),
    Phylum = if_else(is.na(Phylum) | Phylum == "", "Unassigned", Phylum),
    Phylum_plot = if_else(OTU %in% top20, Phylum, "Other")
  ) %>%
  group_by(Sample, Number, Phylum_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Put "Other" at the end of the legend
df$Phylum_plot <- factor(
  df$Phylum_plot,
  levels = c(setdiff(unique(df$Phylum_plot), "Other"), "Other")
)

# 5) Colors: 20 phyla + Other
other_color <- "#C0C0C0"
plot_colors <- c(my_colors[1:20], other_color)

# Optional: make sure color vector matches factor levels exactly
levels_phyla <- levels(df$Phylum_plot)
names(plot_colors) <- levels_phyla  # map colors to labels

# 6) Plot
ggplot(df, aes(x = Number, y = Abundance, fill = Phylum_plot)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Bacteria/Archaea Phyla (Top 20 + Other)",
    x = "Sample Number",
    y = "Relative abundance (%)",
    fill = "Phylum"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 5))   # increase if still crowded


```

::: {.callout-note}
### 💬 Discussion — Phyla composition

The bacterial community across samples is dominated by **Pseudomonadota** (Proteobacteria; which are fast-growing, opportunistic bacteria; copiotrophs), with consistent contributions from **Bacteroidota** and **Actinomycetota**. Minor but **detectable archaeal phyla** were present at low relative abundance. Overall community structure is broadly similar across samples.

Note: Pseudomonadota + Bacteroidota dominance together is a classic signature of active carbon processing systems (refs)
:::

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# Collapse to Class level
ps_class <- tax_glom(ps_16s, taxrank = "Class")

#  Transform to relative abundance
ps_rel <- transform(ps_class, "compositional")

#  Get top 10 classes (by mean relative abundance)
top10 <- top_taxa(ps_rel, n = 10)

#  Melt full dataset
df <- psmelt(ps_rel)
df$Number <- as.factor(df$Number)

# Create Class_plot column (Top 10 + Other)
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Sample, Number, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Put "Other" at the end of legend
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# Adding the "other" colour to the last 10 colours of me palette
other_color <- "#3c3c3c"
plot_colors <- c(my_colors[11:20], other_color)
names(plot_colors) <- levels(df$Class_plot)

# Plot
ggplot(df, aes(x = Number, y = Abundance, fill = Class_plot)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colors) + 
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 10 Classes + Other",
    x = "Sample Number",
    y = "Relative Abundance (%)",
    fill = "Class"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))
```

### Plotting by fjords and depth
#### Sermilik

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=10}

# Collapse + compositional
ps_class <- tax_glom(ps_16s, taxrank = "Class")
ps_rel   <- transform(ps_class, "compositional")

# Top 10 taxa (OTU/ASV IDs after glomming)
top10 <- top_taxa(ps_rel, n = 10)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice (adapt if your Depth column differs)
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(Station)))))
  )

# Top10 + Other
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Class_plot, Water_mass) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# Adding the "other" colour to the last 10 colours of me palette
other_color <- "#3c3c3c"
plot_colors <- c(my_colors[11:20], other_color)
names(plot_colors) <- levels(df$Class_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::label_number(scale = 100, accuracy = 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1),
    expand = c(0, 0)
    ) +
    labs(
      title = paste0(fjord_name, " — Top 10 Classes + Other"),
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 3, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

# Print one (e.g., Sermilik)
plots[["Sermilik"]]

```

#### Iterdalk

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=10}
plots[["Iterdalk"]]

```

#### Tunulliarfik

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=10}
plots[["Tunulliarfik"]]

```

#### Combined

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# Adding the "other" colour to the last 10 colours of me palette
other_color <- "#3c3c3c"
plot_colors <- c(my_colors[11:20], other_color)
names(plot_colors) <- levels(df$Class_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::label_number(scale = 100, accuracy = 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1),
    expand = c(0, 0)
    ) +
    labs(
      title = fjord_name,
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "none"
    ) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

### Plotting by water masses

#### Metling water

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

plot_water_mass <- function(df, water_mass){
  dff <- df %>% filter(Water_mass == water_mass)

  ggplot(dff, aes(x = Sample, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.85) +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = paste0(water_mass, " — Top 10 Classes + Other"),
      x = "Sample name",
      y = "Relative abundance",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(face = "bold")
    ) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
}

WM <- sort(unique(df$Water_mass))

plots <- lapply(WM, function(fj) plot_water_mass(df, fj))
names(plots) <- as.character(WM)

# Print one (e.g., Melting water)
plots[["MW"]]

```

#### Atlantic Water

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots[["AW"]]

```

#### West Greenland Shelf Water

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots[["WGSW"]]

```

#### West Greenland Irminger Water

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots[["WGIW"]]

```

## Targeted ecological follow-up

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

taxa_interest <- c("Thermoplasmata", "Nitrososphaeria","Actinobacteria")

df_interest <- df %>%
  filter(Class_plot %in% taxa_interest)

plot_fjord_interest <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(position = "stack", width = 0.85) +
    facet_wrap(~Depth, nrow = 1) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = paste0(fjord_name, " 16S — Relative abundance of selected classes"),
      x = "Station Number",
      y = "Relative abundance",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(face = "bold")
    )
}

plots_interest <- lapply(fjords, function(fj) plot_fjord_interest(df_interest, fj))
names(plots_interest) <- as.character(fjords)

plots_interest[["Sermilik"]]

```
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots_interest[["Iterdalk"]]
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots_interest[["Tunulliarfik"]]
```

::: {.callout-note}
### 💬 Discussion — Archaeal distributions

Archaeal communities showed a **consistent vertical structure** across all three fjords, with both **Nitrososphaeria** and **Thermoplasmata** largely absent from surface waters and increasing in abundance with depth. This pattern indicates that archaeal activity is concentrated in subsurface and deeper layers rather than in the surface production zone.

**Nitrososphaeria** (likely **Candidatus Nitrosopumilus**) exhibited a strong depth-associated increase, particularly at 45 m, consistent with their role as **ammonia-oxidizing archaea** involved in **nitrification**. Their distribution suggests that nitrification primarily occurs in deeper, low-light waters where regenerated ammonium becomes available. In Tunulliarfik, their presence at 15 m indicates a shallower nitrification zone compared to the other fjords.

**Thermoplasmata** (Marine Group II) were generally more abundant and displayed a broader vertical distribution, with pronounced peaks at depth. This pattern is consistent with their role as **heterotrophic archaea** involved in **organic matter degradation**. The particularly high abundance observed in Iterdalik suggests enhanced carbon processing in this fjord.

The co-occurrence of both groups at depth across all fjords suggests a **tight coupling between carbon remineralization and nitrogen cycling**, whereby organic matter degradation by Thermoplasmata could be fueling ammonium production that is subsequently oxidized by Nitrososphaeria.

**Actinobacteria**

The Actinobacteria are dominated by **Micrococcales** which are typically **aerobic, heterotrophic, and metabolically versatile**. They are well known for their ability to utilise a wide range of organic compounds, including relatively **low concentrations of dissolved organic matter**, and are often considered oligotrophic or stress-tolerant generalists. Many members of this order are also **psychrotolerant**, allowing them to persist in cold environments, and are capable of coping with environmental stressors such as variable salinity, high turbidity, and oxidative stress. They were more abundant in surface waters of the inner regions of Sermilik and Tunulliarfik fjords, both influenced by marine-terminating glaciers. This pattern is consistent with strong glacier-driven environmental gradients.

:::


### Investigating Thermoplasmata and Nitrososphaeria relationship

Thermoplasmata and Nitrososphaeria seem to increase and decrease together. At first glance, this could just be coincidence — or simply the result of both groups preferring the same environmental conditions, like deeper waters.

let's test is this relationship is real:

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# Transform to CLR
ps_clr <- transform(ps_16s, "clr")

df <- psmelt(ps_clr)

df_sub <- df %>%
  filter(Class %in% c("Thermoplasmata", "Nitrososphaeria")) %>%
  group_by(Sample, Depth, Fjord, Class) %>%   # 👈 include metadata HERE
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  pivot_wider(
    names_from = Class,
    values_from = Abundance,
    values_fill = 0
  )

# Correlation
cor.test(df_sub$Thermoplasmata, df_sub$Nitrososphaeria)

# Regression
lm(Thermoplasmata ~ Nitrososphaeria, data = df_sub)

```

With r = 0.72 → strong positive correlation and
p ≈ 1.3 × 10⁻¹¹ → highly significant,
we can say that :"When Nitrososphaeria increases, Thermoplasmata also tends to increase"

The regression tells us that: Thermoplasmata = 10.31 + 1.24 × Nitrososphaeria
For every unit increase in Nitrososphaeria, Thermoplasmata increases by ~1.24

Is this a real biological relationship, or just a shared response to environment (e.g. depth)?
Let's improve the model to include depth and fjord:

```{r, message=FALSE, warning=FALSE}
m2 <- lm(Thermoplasmata ~ Nitrososphaeria + Depth + Fjord, data = df_sub)
summary(m2)
```

The relationship between Thermoplasmata and Nitrososphaeria is real, not just a depth effect!

Thermoplasmata and Nitrososphaeria exhibited a strong positive relationship (p < 0.001), which remained significant after accounting for depth and fjord effects. This suggests potential coupling between archaeal organic matter processing and nitrification processes in glacier-influenced fjord waters.

*Note*: the fjord effect is telling us that: Thermoplasmata is much lower in Sermilik and Tunulliarfik than in Iterdalik, suggesting environmental differences... it makes sense as Iterdalik is a land-termoianted fjord!

There are a couple of possible explanations:

- They may share a similar ecological niche, thriving under the same low-energy, deeper-water conditions.

- Or they could be functionally connected. For example:

  - Nitrososphaeria are known for their role in nitrification (oxidising ammonia)
  - Thermoplasmata may be involved in organic matter degradation

