# 18S Dataset

## 📦 Build the phyloseq object (18S)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/18S"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Make sure taxonomy is a proper matrix and has consistent column names
taxa <- as.matrix(taxa)
target_cols <- c(
  "Kingdom","Subkingdom_1","Subkingdom_2","Phylum","Subphylum","Subphylum_1","Superclass",
  "Class","Subclass","Infraclass","Superorder","Order","Suborder_1","Suborder_2","Suborder_3",
  "Family","Subfamily","Genus","Subgenus","Species"
)
colnames(taxa) <- target_cols[seq_len(ncol(taxa))]

# Load metadata
metadat <- readxl::read_xlsx(file.path(datadir, "metadata_18S.xlsx")) %>%
  mutate(
    Date = as.Date(as.numeric(Date), origin = "1899-12-30"),
    Time = as.POSIXct(as.numeric(Time) * 86400, origin = "1970-01-01", tz = "UTC"),
    Time = format(Time, "%H:%M:%S")
  ) %>% as.data.frame()

rownames(metadat) <- rownames(seqtab.nochim)

# Create a Phyloseq object

ps_18s <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))

cat("🔹 A summary of the phyloseq object \n")
ps_18s

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_18s))
names(dna) <- taxa_names(ps_18s)
ps_18s <- merge_phyloseq(ps_18s, dna)
taxa_names(ps_18s) <- paste0("ASV", seq(ntaxa(ps_18s)))

```
```{r, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 These are the metadata variables:\n")
print(sample_variables(ps_18s))
```
### Positive and negative control

:::: {.columns}

::: {.column width="40%"}
```{r, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

ps_controls <- subset_samples(
  ps_18s,
  Name %in% c("65_Positive_control", "66_Extraction_negative")
)

ps_controls <- prune_taxa(taxa_sums(ps_controls) > 0, ps_controls)

kable(sample_sums(ps_controls), caption = "Number of reads in the controls") %>% kable_styling(full_width = FALSE, position = "center")
```
:::

::: {.column width="20%"}
:::

::: {.column width="40%"}
```{r, message=FALSE, warning=FALSE}

kable(estimate_richness(ps_controls, measures = "Observed"), caption = "Observed richness") %>% kable_styling(full_width = FALSE, position = "center")
```
:::
::::

::: {.callout-note}
### 💬 Discussion — Controls

Those control are less reassuring than the 16S ones:

**Positive control**: 438k reads, 171 ASVs → 171 ASVs sounds high, but can have different explanations!

**Extraction negative**: Extraction negative: 243 reads, 3 ASVs → This is still very low, but higher than the 16S negative (which had 4 reads).

Let's dive into it!
:::

#### Negative control
:::{.column width="80%"}
```{r, message=FALSE, warning=FALSE}
df_neg <- psmelt(ps_controls) %>%
  filter(Sample == "66_Extraction_negative") %>%
  group_by(Genus, Family, Phylum) %>%
  summarise(Reads = sum(Abundance)) %>%
  arrange(desc(Reads))

df_neg %>% head(2) %>% kable()

```
:::

::: {.callout-note}
### 💬 Discussion — Negative control

The 18S extraction negative control contained 243 reads distributed across **two dinoflagellate genera** that (as we will see later) are also abundant in the environmental samples. Given the low read count (~0.1% of a average sample depth) and taxonomic identity, these sequences likely reflect minimal **index cross-talk** rather than true contamination. No need for decontamination!
:::

#### Positive control

:::{.column width="60%"}
```{r, message=FALSE, warning=FALSE}
df_pos <- psmelt(ps_controls) %>%
  filter(Sample == "65_Positive_control")

df_pos %>%
     group_by(OTU) %>%
     summarise(Reads = sum(Abundance)) %>%
     arrange(desc(Reads)) %>%
     slice(1:10) %>% kable()

```
:::

::: {.callout-note}
### 💬 Discussion — Positive control

From the Zymo standard, we are expecting 2 yeasts:

🍄 Yeasts:

 - *Saccharomyces cerevisiae*
 - *Cryptococcus neoformans*

The 18S positive control was dominated by the expected yeast taxa (**Saccharomyces cerevisiae** and **Cryptococcus spp.**), confirmed via BLAST analysis (the MZG database is not adapted in this case).

However, several marine dinoflagellate ASVs were also detected at low relative abundance (~1–2% each). Given the dominance of yeast sequences and the minimal read count in the extraction negative control, these minor marine taxa most likely reflect low-level cross-contamination during PCR or library preparation rather than sample mislabeling or widespread contamination. The overall composition confirms successful amplification of the mock community while indicating minor carryover from environmental samples.

We can now remove the controls from the phyloseq object and move on with the ecological interpretation.
:::

```{r, message=FALSE, warning=FALSE}
# Environmental samples only
ps_18s <- subset_samples(
  ps_18s,
  !Name %in% c("65_Positive_control", "66_Extraction_negative")
)

# Remove zero taxa
ps_18s <- prune_taxa(taxa_sums(ps_18s) > 0, ps_18s)

# saveRDS(ps_18s, "data/ps_18s.rds")

```

## 📊 Data visualisation

In this section, we explore the community structure of the RoCSI 18S dataset using the
`phyloseq` object (`ps_18s`) generated above. We start by visualising read depth
per sample, followed by basic taxonomic composition summaries.

### Read depth per sample
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
read_depth <- data.frame(
Sample = sample_names(ps_18s),
Reads = sample_sums(ps_18s),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() + scale_y_continuous(labels = scales::label_number()) +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "Sequencing depth per sample",
x = "Sample",
y = "Number of non-chimeric reads"
)
```

### Alpha diversity

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}


# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_18s)$Fjord <- as.factor(sample_data(ps_18s)$Fjord)

plot_richness(
  ps_18s,
  x = "Number",
  measures = c("Observed", "Shannon", "Simpson"),
  color = "Fjord", title = "18S Alpha diversity") +
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 4)]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns

Alpha diversity of 18S eukaryotic communities varied among fjords, with Ocean and Tunulliarfik samples generally exhibiting higher richness compared to Sermilik. Shannon and Simpson indices indicate relatively high evenness across most samples, although occasional dominance events were observed (diversity drops), particularly in Sermilik and Tunulliarfik. Compared to 16S prokaryotic diversity, 18S communities displayed greater variability among fjords, suggesting that eukaryotic plankton communities may be more sensitive to local environmental gradients such as stratification and meltwater influence.  
:::

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}


# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_18s)$Fjord <- as.factor(sample_data(ps_18s)$Fjord)
sample_data(ps_18s)$Depth <- as.factor(sample_data(ps_18s)$Depth)

plot_richness(
  ps_18s,
  x = "Number",
  measures = c("Observed", "Shannon", "Simpson"),
  color = "Depth", title = "18S Alpha diversity") +
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 4)]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns
No obvious trends with depth... Maybe surface water tend to exhibit lower diversity...
:::

### Ordination

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
ord <- ordinate(ps_18s, method = "PCoA", distance = "bray")

# Create the ordination plot with ellipses
plot_ordination(ps_18s, ord, color = "Fjord") +
  geom_point(size = 3) +
  ggtitle("Ordination – 18S") +
  stat_ellipse(aes(group = Fjord), type = "norm", linetype = "dashed", size = 1) +
  theme_minimal()
```
::: {.callout-note}
### 💬 Discussion — Beta diversity

The 18S ordination reveals partial clustering by fjord but with substantial overlap among systems. Compared to the 16S bacterial communities, eukaryotic assemblages show greater within-fjord variability and weaker separation among fjords. Tunulliarfik displays the highest dispersion, consistent with elevated alpha diversity and environmental heterogeneity. Overall, these results suggest that while fjord identity influences eukaryotic community structure, planktonic assemblages may be shaped more strongly by shared regional drivers such as stratification and bloom dynamics than by fjord-specific environmental filtering alone.
:::

::: {style="width:60%; margin:auto;"}
```{r, message=FALSE, warning=FALSE}
# Extract sample coordinates from the PCoA result
site <- as.data.frame(ord$vectors)
site$SampleID <- rownames(site)

# Get the 5 lowest values on Axis.1 (side of plot)
bottom5 <- site[order(site$Axis.1), ][1:5, c("SampleID", "Axis.1", "Axis.2")]
kable(bottom5) %>% kable_styling(full_width = FALSE, position = "center")
```
:::

::: {.callout-note}
Again, the most divergent samples in the ordination correspond exclusively to surface (D1) samples. Particularly they are the stations surrounded by icebergs (Fjord Sermilik and Tunulliarfik). These stations were influenced by meltwater and low salinity conditions, suggesting that surface stratification and freshwater input may exert a strong effect on eukaryotic community composition.
:::

### Proportion of unassigned ASVs

```{r, message=FALSE, warning=FALSE}
tax_df <- as.data.frame(tax_table(ps_18s))

unclassified_summary <- tax_df %>%
  summarise(across(
    everything(),
    ~ sum(
      is.na(.) |
      . == "" |
      . == "Unclassified"
    )
  )) %>%
  pivot_longer(
    everything(),
    names_to = "Rank",
    values_to = "n_unclassified_ASVs"
  ) %>%
  mutate(
    percent_unclassified = round(
      100 * n_unclassified_ASVs / ntaxa(ps_18s),
      1
    )
  )

unclassified_summary
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

unclassified_summary <- unclassified_summary %>%
  mutate(
    Rank = factor(Rank, levels = target_cols)
  )

ggplot(unclassified_summary,
       aes(x = Rank, y = percent_unclassified)) +
  geom_col() +
  theme_classic() +
  labs(
    x = "Taxonomic rank",
    y = "% unclassified ASVs"
  )+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

Maybe I should use a different database than MZG for assigning taxonomy...? Would the number of unassigned ASVs decrease?

### Classifing with PR2

```{r, message=FALSE, warning=FALSE}
# library(dada2)
# 
# seqtab.nochim <- readRDS(file.path("data/18S", "seqtab_nochim.rds"))
# 
# taxa_pr2 <- assignTaxonomy(
#   seqs = seqtab.nochim,
#   refFasta = "../pr2_version_5.1.1_SSU_dada2.fasta.gz",
#   multithread = TRUE,
#   minBoot = 50,
#   taxLevels = c(
#     "Domain",
#     "Supergroup",
#     "Division",
#     "Subdivision",
#     "Class",
#     "Order",
#     "Family",
#     "Genus",
#     "Species"
#   )
# )

taxa_pr2 <- readRDS("data/18S/taxa_pr2.rds")
seqtab.nochim <- readRDS("data/18S/seqtab_nochim.rds")

# Load metadata
metadat <- readxl::read_xlsx("data/18S/metadata_18S.xlsx") %>%
  mutate(
    Date = as.Date(as.numeric(Date), origin = "1899-12-30"),
    Time = as.POSIXct(as.numeric(Time) * 86400, origin = "1970-01-01", tz = "UTC"),
    Time = format(Time, "%H:%M:%S")
  ) %>% as.data.frame()

rownames(metadat) <- rownames(seqtab.nochim)

# Create a Phyloseq object

ps_18s_pr2 <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa_pr2))

cat("🔹 A summary of the phyloseq object \n")
ps_18s_pr2

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_18s_pr2))
names(dna) <- taxa_names(ps_18s_pr2)
ps_18s_pr2 <- merge_phyloseq(ps_18s_pr2, dna)
taxa_names(ps_18s_pr2) <- paste0("ASV", seq(ntaxa(ps_18s_pr2)))
```

```{r, message=FALSE, warning=FALSE}
# Environmental samples only
ps_18s_pr2 <- subset_samples(
  ps_18s_pr2,
  !Name %in% c("65_Positive_control", "66_Extraction_negative")
)

# Remove zero taxa
ps_18s_pr2 <- prune_taxa(taxa_sums(ps_18s_pr2) > 0, ps_18s_pr2)

saveRDS(ps_18s_pr2, "data/ps_18s_pr2.rds")
```

```{r, message=FALSE, warning=FALSE}
tax_df <- as.data.frame(tax_table(ps_18s_pr2))

unclassified_summary <- tax_df %>%
  summarise(across(
    everything(),
    ~ sum(
      is.na(.) |
      . == "" |
      . == "Unclassified"
    )
  )) %>%
  pivot_longer(
    everything(),
    names_to = "Rank",
    values_to = "n_unclassified_ASVs"
  ) %>%
  mutate(
    percent_unclassified = round(
      100 * n_unclassified_ASVs / ntaxa(ps_18s),
      1
    )
  )

unclassified_summary
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

taxLevels = c("Domain", "Supergroup","Division", "Subdivision","Class",
              "Order","Family", "Genus","Species")
  
unclassified_summary <- unclassified_summary %>%
  mutate(
    Rank = factor(Rank, levels = taxLevels)
  )

ggplot(unclassified_summary,
       aes(x = Rank, y = percent_unclassified)) +
  geom_col() +
  theme_classic() +
  labs(
    x = "Taxonomic rank",
    y = "% unclassified ASVs"
  )+
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

Ok so from now on, ps_18S is replaced by ps_18S_PR2 as it has far less unassigned ASVs! Eventually, later we can complement the assignment with the MZG database.

```{r, message=FALSE, warning=FALSE}
ps_18s <- ps_18s_pr2
```

### Phylum composition

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",
               "#984EA3", "#A6D854", "#FFFF33", "#FF7F00", "#F781BF", "#999999",                  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F",
               "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")

library(microbiome)

# 1) Agglomerate to Phylum
ps_phylum <- tax_glom(ps_18s, taxrank = "Subdivision")

# 2) Transform to relative abundance (compositional)
ps_rel <- microbiome::transform(ps_phylum, "compositional")

# 3) Top 20 taxa by mean relative abundance (returns taxa_names / OTU IDs)
top20 <- microbiome::top_taxa(ps_rel, n = 20)

# 4) Melt full object
df <- psmelt(ps_rel)

# Ensure Phylum label exists and is character (handles NAs)
df <- df %>%
  mutate(
    Subdivision = as.character(Subdivision),
    Subdivision = if_else(is.na(Subdivision) | Subdivision == "", "Unassigned", Subdivision),
    Subdivision_plot = if_else(OTU %in% top20, Subdivision, "Other")
  ) %>%
  group_by(Sample, Number, Subdivision_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Put "Other" at the end of the legend
df$Subdivision_plot <- factor(
  df$Subdivision_plot,
  levels = c(setdiff(unique(df$Subdivision_plot), "Other"), "Other")
)

# 5) Colors: 20 phyla + Other
other_color <- "#C0C0C0"
plot_colors <- c(my_colors[1:20], other_color)

# Optional: make sure color vector matches factor levels exactly
levels_Subdivision <- levels(df$Subdivision_plot)
names(plot_colors) <- levels_Subdivision  # map colors to labels

# 6) Plot
ggplot(df, aes(x = Number, y = Abundance, fill = Subdivision_plot)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Eukaryote Subdivision (Top 20 + Other)",
    x = "Sample Number",
    y = "Relative abundance (%)",
    fill = "Subdivision"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 4))   # increase if still crowded

```

::: {.callout-note}
### 💬 Discussion — Phyla composition

At the Phylum level, eukaryotic communities are strongly dominated by **Myzozoa**, likely **dinoflagellate lineages**, across most samples. **Bacillariophyta (diatoms)** and **Chlorophyta (green algae)** contribute variably among samples, reflecting potential shifts in primary producer composition.
:::

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# 1️⃣ Collapse to Class level
ps_Class <- tax_glom(ps_18s, taxrank = "Class")

# 2️⃣ Transform to relative abundance
ps_rel <- transform(ps_Class, "compositional")

# 3️⃣ Get top 10 classes (by mean relative abundance)
top10 <- top_taxa(ps_rel, n = 10)

# 4️⃣ Melt full dataset
df <- psmelt(ps_rel)
df$Number <- as.factor(df$Number)

# 5️⃣ Create Class_plot column (Top 10 + Other)
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Sample, Number, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# 6️⃣ Put "Other" at the end of legend
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# 7️⃣ Use your my_colors (first 10) + grey for Other
other_color <- "#C0C0C0"
plot_colors <- c(my_colors[1:10], other_color)
names(plot_colors) <- levels(df$Class_plot)

# 8️⃣ Plot
ggplot(df, aes(x = Number, y = Abundance, fill = Class_plot)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colors) + 
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 10 classes + Other",
    x = "Sample Number",
    y = "Relative Abundance (%)",
    fill = "Class"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5),
    breaks  = function(x) x[seq(1, length(x), by = 4)]
  )
```
::: {.callout-note}
### 💬 Discussion — Class composition

At the Class level, eukaryotic communities were strongly dominated by **Dinophyceae (dinoflagellates)** across most samples, with episodic increases in **Bacillariophyceae (diatoms) and copepods**. These patterns suggest that **dinoflagellates** and associated trophic interactions **play a central role** in structuring the planktonic community. Compared to bacterial communities, 18S profiles exhibit stronger dominance patterns and greater patchiness.

**NOTE**: Dinoflagellates can have extremely high numbers of 18S rRNA gene copies per cell (~ 1000 to ~ 10.000)!
--> One dinoflagellate cell can produce the same number of reads as hundreds or thousands of other cells

Better to say: likely dominant, but their relative abundance is exaggerated by molecular biases.

**NOTE**: Mamiellophyceae are often the numerically dominant group but they have low 18S copy number tendency and are likely under-represented!
:::


### Plotting by fjords and depth
#### Sermilik

```{r, message=FALSE, warning=FALSE, fig.width=8.5, fig.height=10}

# Collapse + compositional
ps_Class <- tax_glom(ps_18s, taxrank = "Class")
ps_rel   <- transform(ps_Class, "compositional")

# Top 10 taxa (OTU/ASV IDs after glomming)
top10 <- top_taxa(ps_rel, n = 10)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice (adapt if your Depth column differs)
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(Station)))))
  )

# Top10 + Other
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# Colors: use your palette (first 10) + Other
other_color <- "#C0C0C0"
plot_colors <- c(my_colors[1:10], other_color)
names(plot_colors) <- levels(df$Class_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::label_number(scale = 100, accuracy = 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1),
    expand = c(0, 0)
    ) +
    labs(
      title = paste0(fjord_name, " — Top 10 Classes + Other"),
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 3, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

# Print one (e.g., Sermilik)
plots[["Sermilik"]]

```

#### Iterdalk

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots[["Iterdalk"]]

```

#### Tunulliarfik

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plots[["Tunulliarfik"]]

```

#### Combined

```{r, message=FALSE, warning=FALSE, fig.width=11, fig.height=8}

plot_colors <- c(
  "Bacillariophyceae" = "#0b6e1a",   # diatoms -> dark green
  "Dinophyceae"       = "#2B55DB",   # dinoflagellates -> blue
  "Arthropoda"          = "#FF4343",   # crustacea / copepods -> red
  "Cryptophyceae"     = "#984EA3",   # flagellates -> purple
  "Oligotrichea"      = "#f0e442",   # ciliates -> light grey

  # no clear direct match in the reference, so choose distinct extras
  "Choanoflagellatea" = "#8dd3c7",
  "Chrysophyceae"     = "#FF7F00",
  "Mamiellophyceae"   = "#d9d9d9",
  "Pyramimonadophyceae" = "#A6D854",
  "Protozoa_incertae_sedis_EXT" = "#e377c2",
  "Other"             = "#3c3c3c"
)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::label_number(scale = 100, accuracy = 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1),
    expand = c(0, 0)
    ) +
    labs(
      title = fjord_name,
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "none"
    ) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

::: {.callout-note}
### 💬 Discussion — Class composition by depth and fjords

**NOTE** We should probably use 18S gene copy number (GCN) correction factors for marine eukaryotes to correct this biases of 18S GCN!

REf: Martin et al. (2022): "Towards quantitative metabarcoding of eukaryotic plankton: an approach to improve 18S rRNA gene copy number bias"
https://doi.org/10.3897/mbmg.6.85794
:::

### GCN-adjusted compositional plot

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# Make a class-level GCN table
gcn_table <- tibble(
  Class = c(
    "Dinophyceae","Oligotrichea","Bacillariophyceae",
    "Cryptophyceae","Mamiellophyceae","Pyramimonadophyceae",
    "Chrysophyceae","Choanoflagellatea","Copepoda",
    "Protozoa incertae sedis","Other / Unclassified"
  ),
    GCN = c(5000, 20000, 20, 300, 4, 10,  100, 50,
    500, 1000, 1 ),
    Evidence = c("High","High","High","Medium","High","Medium",
    "Medium","Low","Low","Low","NA"),
  
  Reference = c("Gong & Marchetti 2019; Galluzzi et al. 2010",
    "Lapeyra Martin et al. 2022","Gong & Marchetti 2019",
    "Lapeyra Martin et al. 2022", "Gong & Marchetti 2019",
    "Gong & Marchetti 2019 (proxy)","Lapeyra Martin et al. 2022 (proxy)",
    "Proxy estimate", "Proxy estimate",
    "Proxy estimate", "Not corrected"
  )
)

library(gt)
gcn_table %>%
  gt() %>%
  tab_header(
    title = "Class-level 18S rRNA Gene Copy Number Corrections",
  ) %>%
  cols_label(
    Class = "Taxonomic Class",
    GCN = "GCN Value",
    Evidence = "Confidence",
    Reference = "Reference"
  ) %>%
  fmt_number(
    columns = GCN,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) %>% 
      data_color(
    columns = Evidence,
    colors = c(
      "High" = "darkgreen",
      "Medium" = "orange",
      "Low" = "red"
    )
  )

```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# Create a GCN-corrected phyloseq object at Class level
library(phyloseq)
library(dplyr)
library(forcats)

# 1) Class-level counts
df_gcn <- ps_18s %>%
  tax_glom(taxrank = "Class", NArm = FALSE) %>%
  psmelt() %>%
  mutate(
    Class = if_else(is.na(Class) | Class == "", "Unclassified", as.character(Class)),
    Depth = case_when(
      Depth %in% c("1","01","surface","Surface") ~ "surface",
      Depth %in% c("15","15m","15 m")            ~ "15m",
      Depth %in% c("45","45m","45 m")            ~ "45m",
      TRUE ~ as.character(Depth)
    ),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(as.character(Station))))))
  ) %>%
  left_join(gcn_table, by = "Class") %>%
  mutate(
    GCN = coalesce(GCN, 1),
    Abundance = Abundance / GCN
  )

# 2) Top 10 classes after GCN correction
top10_classes <- df_gcn %>%
  group_by(Class) %>%
  summarise(total = sum(Abundance), .groups = "drop") %>%
  slice_max(total, n = 10) %>%
  pull(Class)

# 3) Top 10 + Other, then compositional within sample
df_gcn <- df_gcn %>%
  mutate(Class_plot = if_else(Class %in% top10_classes, Class, "Other")) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Fjord, Depth, Sample, Number, Station) %>%
  mutate(Abundance = Abundance / sum(Abundance)) %>%
  ungroup()

# 4) Legend order
df_gcn$Class_plot <- factor(
  df_gcn$Class_plot,
  levels = c(setdiff(sort(unique(df_gcn$Class_plot)), "Other"), "Other")
)

plot_fjord <- function(df_gcn, fjord_name){
  dff <- df_gcn %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(labels = scales::label_number(scale = 100, accuracy = 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1),
    expand = c(0, 0)
    ) +
    labs(
      title = fjord_name,
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "none"
    ) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
}

fjords <- sort(unique(df_gcn$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df_gcn, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

## Targeted ecological follow-up

### Looking at some taxa of interest

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

taxa_interest <- c("Mamiellophyceae", "Oligotrichea","Choanoflagellatea")

df_interest <- df %>%
  filter(Class_plot %in% taxa_interest)

plot_fjord_interest <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Class_plot)) +
    geom_col(position = "stack", width = 0.85) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = fjord_name,
      x = "Station Number",
      y = "Relative abundance",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(face = "bold")
    )
}

plots_interest <- lapply(fjords, function(fj) plot_fjord_interest(df_interest, fj))
names(plots_interest) <- as.character(fjords)

plots_interest[["Tunulliarfik"]] <- plots_interest[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots_interest[["Iterdalk"]] <- plots_interest[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots_interest[["Sermilik"]] + 
plots_interest[["Tunulliarfik"]] + 
plots_interest[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

### Looking at the Dinoflagelates and Diatoms communities

#### Plotting at Species level

##### Dinoflagellates

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(microbiome)

my_colors <- c("#FF6666", "#8DD3C7", "#377EB8","#5DA5A4" , "#984EA3", "#A6D854", "#FFFF33",
               "#FF7F00", "#F781BF", "#66C2A5","#4DAF4A", "#FC8D62", "#999999", "#8DA0CB", "#E78AC3", 
               "#FFD92F", "#E5C494", "#B3B3B3", "#A65628", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")


# Keep only Bacillariophyceae and Dinophyceae
ps_bd <- subset_taxa(
  ps_18s,
  Class %in% c("Dinophyceae") # %in% c("Bacillariophyceae", "Dinophyceae")
)

# Collapse to species level
ps_species <- tax_glom(ps_bd, taxrank = "Genus", NArm = FALSE)

# Convert to relative abundance
# This makes abundances relative to the filtered community only:
# Bacillariophyceae + Dinophyceae = 100%
ps_rel <- transform_sample_counts(ps_species, function(x) x / sum(x))

# Top 10 taxa/genus-level OTU IDs after glomming
top10 <- top_taxa(ps_rel, n = 15)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice + clean species names
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),

    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = rev(as.character(sort(unique(as.numeric(Station)))))),

    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified class", Class),

    Genus = as.character(Genus),
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified genus", Genus),

    Genus_label =  Genus
  )

# Top15 species + Other
df <- df %>%
  mutate(
    Genus_plot = if_else(OTU %in% top10, Genus_label, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Genus_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Genus_plot <- factor(
  df$Genus_plot,
  levels = c(setdiff(sort(unique(df$Genus_plot)), "Other"), "Other")
)

other_color <- "black"

plot_colors <- c(
  rep(my_colors, length.out = length(levels(df$Genus_plot)) - 1),
  other_color
)

names(plot_colors) <- levels(df$Genus_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Genus_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    labs(
      title = fjord_name,
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Genus"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 3, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

##### Diatoms

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(microbiome)

my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A", "#984EA3", "#A6D854", "#FFFF33",
               "#FF7F00", "#F781BF", "#999999", "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", 
               "#FFD92F", "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")


# Keep only Bacillariophyceae and Dinophyceae
ps_bd <- subset_taxa(
  ps_18s,
  Class %in% c("Bacillariophyceae") # %in% c("Bacillariophyceae", "Dinophyceae")
)

# Collapse to species level
ps_species <- tax_glom(ps_bd, taxrank = "Species", NArm = FALSE)

# Convert to relative abundance
# This makes abundances relative to the filtered community only:
# Bacillariophyceae + Dinophyceae = 100%
ps_rel <- transform_sample_counts(ps_species, function(x) x / sum(x))

# Top 10 taxa/genus-level OTU IDs after glomming
top10 <- top_taxa(ps_rel, n = 15)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice + clean species names
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),

    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(Station))))),

    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified class", Class),

    Genus = as.character(Genus),
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified genus", Genus),

    Species = as.character(Species),
    Species = if_else(is.na(Species) | Species == "", "Unclassified species", Species),

    Species_label = paste(Class, Genus, Species, sep = " — ")
  )

# Top15 species + Other
df <- df %>%
  mutate(
    Species_plot = if_else(OTU %in% top10, Species_label, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Species_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Species_plot <- factor(
  df$Species_plot,
  levels = c(setdiff(sort(unique(df$Species_plot)), "Other"), "Other")
)

other_color <- "#C0C0C0"

plot_colors <- c(
  rep(my_colors, length.out = length(levels(df$Species_plot)) - 1),
  other_color
)

names(plot_colors) <- levels(df$Species_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Species_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    labs(
      title = paste0(fjord_name, " — Species"),
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Species"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 6, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

#### Plotting at Genus level

##### Dinoflagellates

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(microbiome)

my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A", "#984EA3", "#A6D854", "#FFFF33",
               "#FF7F00", "#F781BF", "#999999", "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", 
               "#FFD92F", "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")


# Keep only Bacillariophyceae and Dinophyceae
ps_bd <- subset_taxa(
  ps_18s,
  Class %in% c("Dinophyceae") # %in% c("Bacillariophyceae", "Dinophyceae")
)

# Collapse to genus level
ps_genus <- tax_glom(ps_bd, taxrank = "Genus", NArm = FALSE)

# Convert to relative abundance
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Top 10 taxa/genus-level OTU IDs after glomming
top10 <- top_taxa(ps_rel, n = 15)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice + clean species names
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),

    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(Station))))),

    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified class", Class),

    Genus = as.character(Genus),
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified genus", Genus),

    Genus_label = paste(Class, Genus, sep = " — ")
  )

# Top15 species + Other
df <- df %>%
  mutate(
    Genus_plot = if_else(OTU %in% top10, Genus_label, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Genus_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Genus_plot <- factor(
  df$Genus_plot,
  levels = c(setdiff(sort(unique(df$Genus_plot)), "Other"), "Other")
)

other_color <- "#C0C0C0"

plot_colors <- c(
  rep(my_colors, length.out = length(levels(df$Genus_plot)) - 1),
  other_color
)

names(plot_colors) <- levels(df$Genus_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Genus_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    labs(
      title = paste0(fjord_name, " — Genus"),
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Genus"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 6, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

##### Diatoms

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(microbiome)

my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A", "#984EA3", "#A6D854", "#FFFF33",
               "#FF7F00", "#F781BF", "#FDB462", "#66C2A5", "#C05A5A", "#8DA0CB", "#999999", 
               "#FFD92F", "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3" )


# Keep only Bacillariophyceae and Dinophyceae
ps_bd <- subset_taxa(
  ps_18s,
  Class %in% c("Bacillariophyceae") # %in% c("Bacillariophyceae", "Dinophyceae")
)

# Collapse to genus level
ps_genus <- tax_glom(ps_bd, taxrank = "Genus", NArm = FALSE)

# Convert to relative abundance
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Top 10 taxa/genus-level OTU IDs after glomming
top10 <- top_taxa(ps_rel, n = 15)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice + clean species names
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),

    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = rev(as.character(sort(unique(as.numeric(Station)))))),

    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified class", Class),

    Genus = as.character(Genus),
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified genus", Genus),

    Genus_label =  Genus
  )

# Top15 species + Other
df <- df %>%
  mutate(
    Genus_plot = if_else(OTU %in% top10, Genus_label, "Other")
  ) %>%
  group_by(Fjord, Depth, Sample, Number, Station, Genus_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Genus_plot <- factor(
  df$Genus_plot,
  levels = c(setdiff(sort(unique(df$Genus_plot)), "Other"), "Other")
)

other_color <- "black"

plot_colors <- c(
  rep(my_colors, length.out = length(levels(df$Genus_plot)) - 1),
  other_color
)

names(plot_colors) <- levels(df$Genus_plot)

plot_fjord <- function(df, fjord_name){
  dff <- df %>% filter(Fjord == fjord_name)

  ggplot(dff, aes(x = Station, y = Abundance, fill = Genus_plot)) +
    geom_col(width = 0.9) +
    facet_grid(rows = vars(Depth), switch = "y") +
    scale_fill_manual(values = plot_colors) +
    scale_y_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    labs(
      title = fjord_name,
      x = "Station Number",
      y = "Relative abundance (%)",
      fill = "Genus"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0, face = "bold", size = 12),
      panel.spacing.y = unit(1, "lines"),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      legend.position = "bottom"
    ) +
    guides(fill = guide_legend(nrow = 3, byrow = TRUE))
}

fjords <- sort(unique(df$Fjord))

plots <- lapply(fjords, function(fj) plot_fjord(df, fj))
names(plots) <- as.character(fjords)

plots[["Tunulliarfik"]] <- plots[["Tunulliarfik"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

plots[["Iterdalk"]] <- plots[["Iterdalk"]] +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.y.left = element_blank())

library(patchwork)

plots[["Sermilik"]] + 
plots[["Tunulliarfik"]] + 
plots[["Iterdalk"]] +
plot_layout(ncol = 3, guides = "collect") & 
  labs(x = "Station Number", y = "Relative abundance (%)") &
  theme(legend.position = "bottom")

```

### Clustering based on environmental conditions

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# Collapse + compositional
ps_class <- tax_glom(ps_18s, taxrank = "Class")
ps_rel   <- transform(ps_class, "compositional")

# Top 10 taxa (OTU/ASV IDs after glomming)
top10 <- top_taxa(ps_rel, n = 10)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(as.character(Station))))))
  )

df <- df %>%
  mutate(
    Sample_cluster = gsub(".*_St([0-9]+)_D([0-9]+)$", "\\1_\\2", Sample)
  )

# Define clusters by Sample names
cluster_list <- list(
  Cluster4 = c("9_1", "10_1", "14_45"),

  Cluster3 = c("1_45", "2_45", "3_45", "4_45", "5_45",
               "8_45", "10_15", "10_45", "11_45", "12_45", "15_1",
               "15_15", "15_45", "16_45", "17_15", "17_45", "18_15", "18_45",
               "19_15", "19_45", "20_15", "20_45", "21_45", "22_45"),

  Cluster2 = c("1_15", "2_15", "3_15", "4_15", "5_1", "5_15", "6_1", "6_15", "6_45",
               "7_1", "7_15", "7_45", "8_1", "8_15", "11_1", "11_15", "12_1",
               "12_15", "13_1", "13_15", "13_45", "14_1", "14_15", "16_1",
               "16_15", "21_15", "22_15"),

  Cluster1 = c("1_1", "2_1", "3_1", "4_1",
               "17_1", "18_1", "19_1", "20_1", "21_1", "22_1")
)

# Add Cluster column using Sample_cluster
df <- df %>%
  mutate(
    Cluster = case_when(
      Sample_cluster  %in% cluster_list$Cluster1 ~ "Cluster1",
      Sample_cluster %in% cluster_list$Cluster2 ~ "Cluster2",
      Sample_cluster %in% cluster_list$Cluster3 ~ "Cluster3",
      Sample_cluster %in% cluster_list$Cluster4 ~ "Cluster4",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Cluster))

# Top10 + Other
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Cluster, Sample_cluster, Depth, Sample, Number, Station, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# Adding the "other" colour to the last 10 colours of my palette
other_color <- "#3c3c3c"
plot_colors <- c(my_colors[11:20], other_color)
names(plot_colors) <- levels(df$Class_plot)

# Optional: order clusters
df$Cluster <- factor(df$Cluster, levels = c("Cluster1", "Cluster2", "Cluster3", "Cluster4"))

plot_cluster <- function(df, cluster_name){
  dff <- df %>% filter(Cluster == cluster_name)

  ggplot(dff, aes(x = Abundance, y = Sample_cluster, fill = Class_plot)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = plot_colors) +
    scale_x_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    labs(
      title = paste0(cluster_name, " — Top 10 Classes + Other"),
      y = "Sample name",
      x = "Relative abundance (%)",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(size = 12),
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      legend.position = "bottom"
    ) 
}

clusters <- levels(df$Cluster)

plots <- lapply(clusters, function(cl) plot_cluster(df, cl))
names(plots) <- as.character(clusters)

# Print one
plots[["Cluster1"]]
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
library(patchwork)

combined_plot <-
  (plots[["Cluster1"]] + plots[["Cluster2"]]) /
  (plots[["Cluster3"]] + plots[["Cluster4"]]) +
  plot_layout(
    guides = "collect",
    widths = c(1, 1),
    heights = c(0.75, 1.25)
  ) &
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    axis.text.y = element_text(size = 6.5),
    axis.text.x = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    panel.spacing = unit(1.2, "lines")
  )

combined_plot
```

### All together

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
plot_colors <- c(
  "Bacillariophyceae" = "#0b6e1a",   # diatoms -> dark green
  "Dinophyceae"       = "#2B55DB",   # dinoflagellates -> blue
  "Copepoda"          = "#FF4343",   # crustacea / copepods -> red
  "Cryptophyceae"     = "#984EA3",   # flagellates -> purple
  "Oligotrichea"      = "#f0e442",   # ciliates -> light grey

  # no clear direct match in the reference, so choose distinct extras
  "Choanoflagellatea" = "#8dd3c7",
  "Chrysophyceae"     = "#FF7F00",
  "Mamiellophyceae"   = "#d9d9d9",
  "Pyramimonadophyceae" = "#A6D854",
  "Protozoa_incertae_sedis_EXT" = "#e377c2",
  "Other"             = "#3c3c3c"
)

# Collapse + compositional
ps_class <- tax_glom(ps_18s, taxrank = "Class")
ps_rel   <- transform(ps_class, "compositional")

# Top 10 taxa (OTU/ASV IDs after glomming)
top10 <- top_taxa(ps_rel, n = 10)

# Melt full object
df <- psmelt(ps_rel)

# Make Depth nice
df <- df %>%
  mutate(
    Depth = as.character(Depth),
    Depth = if_else(Depth %in% c("1","01","surface","Surface"), "surface",
             if_else(Depth %in% c("15","15m","15 m"), "15m",
             if_else(Depth %in% c("45","45m","45 m"), "45m", Depth))),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord),
    Station = factor(Station, levels = as.character(sort(unique(as.numeric(as.character(Station))))))
  )

df <- df %>%
  mutate(
    Sample_cluster = gsub(".*_St([0-9]+)_D([0-9]+)$", "\\1_\\2", Sample)
  )

# Define clusters by Sample names
cluster_list <- list(
  Cluster4 = c("9_1", "10_1", "14_45"),

  Cluster3 = c("1_45" , "2_45" , "3_45"  ,"4_45","8_45" , "10_45" ,"11_45", "12_45",
               "13_45", "15_1" , "15_15" , "15_45", "17_15" ,"17_45" ,"18_15", "18_45",
               "19_15" ,"19_45" ,"20_15" ,"20_45", "21_45", "22_45"),

  Cluster2 = c("1_15" , "2_15" , "3_15",  "4_15" , "5_1"  , "5_15" , "5_45", "6_1" , 
               "6_15" , "6_45" , "7_1" , "7_15" , "7_45" , "8_1"  , "8_15" , "10_15",
               "11_1",  "11_15", "12_1" , "12_15", "13_1",  "13_15" , "14_1" , "14_15" ,
               "16_1" , "16_15" ,"16_45",  "21_15" ,"22_15"),

  Cluster1 = c("1_1", "2_1", "3_1", "4_1", "17_1", "18_1", "19_1", "20_1", "21_1", "22_1")
)

# Add Cluster column using Sample_cluster
df <- df %>%
  mutate(
    Cluster = case_when(
      Sample_cluster  %in% cluster_list$Cluster1 ~ "Cluster1",
      Sample_cluster %in% cluster_list$Cluster2 ~ "Cluster2",
      Sample_cluster %in% cluster_list$Cluster3 ~ "Cluster3",
      Sample_cluster %in% cluster_list$Cluster4 ~ "Cluster4",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Cluster))

sample_order <- c(
  "22_1", "21_1", "20_1", "19_1", "18_1", "17_1", "4_1", "3_1", "2_1", "1_1",
  "22_15", "21_15", "16_45", "16_15", "16_1", "14_15", "14_1", "13_15", "13_1",
  "12_15", "12_1", "11_15", "11_1", "10_15", "8_15", "8_1", "7_45", "7_15",
  "7_1", "6_45", "6_15", "6_1", "5_45", "5_15", "5_1", "4_15", "3_15", "2_15", "1_15",
  "22_45", "21_45", "20_45", "20_15", "19_45", "19_15", "18_45", "18_15",
  "17_45", "15_45", "15_15", "15_1", "13_45", "12_45", "11_45", "10_45",
  "8_45", "4_45", "3_45", "1_45","14_45", "10_1", "9_1"
)

# Top10 + Other
df <- df %>%
  mutate(
    Class = as.character(Class),
    Class = if_else(is.na(Class) | Class == "", "Unclassified", Class),
    Class_plot = if_else(OTU %in% top10, Class, "Other")
  ) %>%
  group_by(Cluster, Sample_cluster, Depth, Sample, Number, Station, Class_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Legend order: Other last
df$Class_plot <- factor(
  df$Class_plot,
  levels = c(setdiff(sort(unique(df$Class_plot)), "Other"), "Other")
)

# Optional: order clusters
df$Cluster <- factor(df$Cluster, levels = c("Cluster1", "Cluster2", "Cluster3", "Cluster4"))

df$Sample_cluster <- factor(df$Sample_cluster, levels = rev(c("1_1" , "2_1" , "3_1" , "4_1" ,
                                           "17_1", "18_1", "19_1", "20_1" ,"21_1", "22_1",
                                           "1_15" , "2_15" , "3_15",  "4_15" , "5_1"  , "5_15" , "6_1" ,  "6_15" , "6_45" ,
                                           "7_1"  , "7_15" , "7_45" , "8_1"  , "8_15" , "11_1",  "11_15", "12_1" ,
                                           "12_15", "13_1",  "13_15" , "13_45", "14_1" ,
                                           "14_15" , "16_1" , "16_15" , "21_15" ,"22_15",
                                           "1_45" , "2_45" , "3_45"  ,"4_45",  "5_45"  ,
                                           "8_45" , "10_15", "10_45" ,"11_45", "12_45", "15_1" ,
                                           "15_15" , "15_45", "16_45" ,"17_15" ,"17_45" ,"18_15", "18_45",
                                           "19_15" ,"19_45" ,"20_15" ,"20_45",
                                           "21_45", "22_45",
                                           "9_1" ,  "10_1" , "14_45")))

ggplot(df, aes(x = Abundance, y = Sample_cluster, fill = Class_plot)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = plot_colors) +
    scale_x_continuous(
      labels = scales::label_number(scale = 100, accuracy = 1),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      expand = expansion(mult = c(0, 0.02))
    ) +
    scale_y_discrete(drop = TRUE) +
    labs(
      title = "Top 10 Classes + Other",
      y = "Sample name",
      x = "Relative abundance (%)",
      fill = "Class"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 11),
      axis.title.y = element_text(size = 14),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      legend.position = "bottom"
    )
```



### Looking at Picozoa distribution

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
library(phyloseq)
library(dplyr)
library(ggplot2)

# Melt full phyloseq object
df_asv43 <- psmelt(ps_18s) %>%
  filter(OTU == "ASV43") %>%
  mutate(
    Depth = case_when(
      Depth %in% c("1","01","surface","Surface") ~ "surface",
      Depth %in% c("15","15m","15 m") ~ "15m",
      Depth %in% c("45","45m","45 m") ~ "45m",
      TRUE ~ as.character(Depth)
    ),
    Depth = factor(Depth, levels = c("surface", "15m", "45m")),
    Fjord = as.factor(Fjord)
  )

df_asv43_sum <- df_asv43 %>%
  group_by(Fjord, Depth) %>%
  summarise(
    mean_reads = mean(Abundance, na.rm = TRUE),
    sd_reads   = sd(Abundance, na.rm = TRUE),
    n          = n(),
    .groups = "drop"
  )

ggplot(df_asv43_sum, aes(x = Depth, y = mean_reads, fill = Fjord)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_reads - sd_reads, ymax = mean_reads + sd_reads),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "ASV43 distribution by fjord and depth",
    x = "Depth",
    y = "Mean read count"
  )

```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
df_asv43_plot <- df_asv43 %>%
  mutate(
    Station = factor(
      Station,
      levels = as.character(sort(unique(as.numeric(as.character(Station)))))
    )
  )

ggplot(df_asv43_plot, aes(x = Station, y = Abundance, fill = Fjord)) +
  geom_col() +
  facet_grid(rows = vars(Depth), cols = vars(Fjord), switch = "y") +
  theme_classic() +
  labs(
    title = "ASV43 across samples",
    x = "Station",
    y = "Read count"
  )
```