# Function prediction - Picrust2

## Running Picrust2

```{r, message=FALSE, warning=FALSE}
library(tidyverse)

# Load DADA2 sequence table
seqtab <- readRDS("data/16S/seqtab_nochim.rds")

# Remove controls
seqtab <- seqtab[!rownames(seqtab) %in% c("65_zymo_positive", "66_Extraction_negative"), ]

# Extract ASV sequences
asv_seqs <- colnames(seqtab)

# Make simple stable ASV IDs
asv_ids <- paste0("ASV", seq_along(asv_seqs))

# Save a lookup table
asv_map <- data.frame(
  ASV = asv_ids,
  Sequence = asv_seqs,
  stringsAsFactors = FALSE
)
write.csv(asv_map, "ASV_map.csv", row.names = FALSE, quote = FALSE)

# Rename seqtab columns from raw sequences to ASV IDs
colnames(seqtab) <- asv_ids

# PICRUSt2 TSV input should have ASVs as rows, samples as columns
picrust_tab <- t(seqtab)

# Write abundance table
write.table(
  picrust_tab,
  file = "picrust2_input.tsv",
  sep = "\t",
  quote = FALSE,
  col.names = NA
)

# Write representative sequences FASTA
con <- file("picrust2_seqs.fna", open = "w")
for (i in seq_along(asv_ids)) {
  writeLines(paste0(">", asv_ids[i]), con)
  writeLines(asv_seqs[i], con)
}
close(con)
```

```bash
picrust2_pipeline.py \
-s picrust2_seqs.fna \
-i picrust2_input.tsv \
-o picrust2_out \
--threads 30 \
--stratified
```
PICRUSt2 is giving you three levels of functional prediction from your 16S data:

 - KO = KEGG Ortholog
 - EC = Enzyme Commission number
 - Pathway = usually MetaCyc pathway in PICRUSt2 by default

They are related, but they are not the same kind of annotation. PICRUSt2 predicts gene-family abundances such as KO and EC, and then uses EC-based mappings to infer higher-level pathway abundances/coverage.

## Pathways: multi-step metabolic processes

A pathway is a higher-level biological process made of multiple enzymatic steps, such as amino-acid biosynthesis, glycolysis-related routes, sulfur metabolism, or degradation pathways. In PICRUSt2, pathway predictions are typically MetaCyc pathways, and they are inferred from the predicted EC content.

```{r, message=FALSE, warning=FALSE}
# If you want to analyze the abundance of KEGG pathways instead of KO within the pathway, please set `ko_to_kegg` to TRUE.
# KEGG pathways typically have more descriptive explanations.

library(readr)
library(ggpicrust2)
library(tibble)
library(tidyverse)
library(ggprism)
library(patchwork)

# Load necessary data: abundance data and metadata
pathway_abun <- read_tsv("data/16S/pathways_out/path_abun_unstrat_descrip.tsv.gz",
                            show_col_types = FALSE)
# keep pathway annotations separately
pathway_info <- pathway_abun %>%
  dplyr::select(pathway, description)

# keep only abundance columns and convert to matrix for downstream analysis
path_mat <- pathway_abun %>%
  dplyr::select(-description) %>%
  column_to_rownames("pathway") %>%
  as.matrix()

metadata <- readxl::read_xlsx("data/16S/metadata_16S.xlsx") %>% 
    filter(!Name %in% c("65_zymo_positive", "66_Extraction_negative")) %>%
  mutate(
    Name = as.character(Name),
    Station = factor(Station),
    Depth = factor(Depth, levels = c("1", "15", "45")),
    Fjord = factor(Fjord),
    Date = as.Date(as.numeric(Date), origin = "1899-12-30"),
    Time = as.POSIXct(as.numeric(Time) * 86400, origin = "1970-01-01", tz = "UTC"),
    Time = format(Time, "%H:%M:%S"),
    Water_mass = factor(Water_mass)
  )

# Normalise abundances
path_rel <- apply(path_mat, 2, function(x) x / sum(x))
```

### Ordination

```{r, message=FALSE, warning=FALSE}
library(vegan)
set.seed(42)

bc <- vegdist(t(path_rel), method = "bray")
ord <- metaMDS(bc, trace = 0)

ord_df <- as.data.frame(ord$points) %>%
  rownames_to_column("Name") %>%
  left_join(metadata, by = "Name")

ggplot(ord_df, aes(MDS1, MDS2, color = Depth)) +
  geom_point(size = 3) +
  theme_bw()
```
```{r, message=FALSE, warning=FALSE}
ggplot(ord_df, aes(MDS1, MDS2, color = Water_mass)) +
  geom_point(size = 3) +
  theme_bw()
```

### Visualise overall pathway differences

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
library(pheatmap)

top_ids <- names(sort(rowMeans(path_mat), decreasing = TRUE))[1:30]

top_mat <- path_rel[top_ids, ]

top_labels <- pathway_info %>%
  filter(pathway %in% top_ids) %>%
  distinct(pathway, description)

# make sure label order matches matrix row order
top_labels <- top_labels %>%
  slice(match(rownames(top_mat), pathway))

rownames(top_mat) <- paste(top_labels$description)

sample_annot <- metadata %>%
  filter(Name %in% colnames(top_mat)) %>%
  column_to_rownames("Name") %>%
  dplyr::select(Depth, Fjord, Water_mass)

ph <- pheatmap(
  top_mat,
  annotation_col = sample_annot,
  fontsize_col = 5,   # smaller column labels
  angle_col = 45      # rotate labels
)

```

Below is the list of the samples that belong to the cluster on the right side of the heatmap figure above: 
```{r, message=FALSE, warning=FALSE}
cl <- cutree(ph$tree_col, k = 2)
ordered_samples <- colnames(top_mat)[ph$tree_col$order]
right_samples_path <- ordered_samples[ordered_samples %in% names(cl[cl == 1])]

right_samples_path
```

These samples are almost exclusively from 1 and 15 m... (there is one sample from 45 m)


## KO: gene-family / ortholog groups

A KO groups homologous genes that are considered to perform an equivalent function in the KEGG system. So a KO is usually the closest of the three outputs to a specific functional gene family. For example, a KO might correspond to a transporter subunit, an enzyme protein, or another functionally defined gene product.

```{r, message=FALSE, warning=FALSE}

# Load necessary data: 
ko <- read_tsv("data/16S/KO_metagenome_out/pred_metagenome_unstrat_descrip.tsv.gz",
                            show_col_types = FALSE) %>% rename(KO_ID = `function`)
# keep pathway annotations separately
ko_info <- ko %>%
  dplyr::select(KO_ID, description)

nitrogen_kos <- c(
  # Nitrogen fixation
  "K02586", # nifH
  "K02588", # nifD
  "K02591", # nifK

  # Ammonium transport / assimilation
  "K03320", # amtB
  "K01915", # glnA

  # Nitrification: ammonia oxidation
  "K10944", # amoA
  "K10945", # amoB
  "K10946", # amoC
  "K10535", # hao

  # Nitrification: nitrite oxidation
  "K00370", # nxrA / narG
  "K00371", # nxrB / narH

  # Membrane-bound nitrate reductase (Nar)
  "K00370", # narG / nxrA
  "K00371", # narH / nxrB
  "K00374", # narI
  "K00373", # narJ

  # Periplasmic nitrate reductase (Nap)
  "K02567", # napA
  "K02568", # napB
  "K02569", # napC
  "K02570", # napD

  # Assimilatory nitrate/nitrite reduction
  "K00372", # nasA
  "K00366", # nirA

  # Denitrification
  "K00368", # nirK
  "K15864", # nirS
  "K04561", # norB
  "K02305", # norC
  "K00376", # nosZ

  # DNRA
  "K03385", # nrfA
  "K15876", # nrfH
  "K00362", # nirB
  "K00363", # nirD

  # Anammox
  "K20932", # hzsA
  "K20933", # hzsB
  "K20934", # hzsC
  "K20935"  # hdh
) |> unique()

ko_n <- ko %>%
  mutate(KO = sub("ko:", "", KO_ID)) %>%
  filter(KO %in% nitrogen_kos)

nitrogen_kos_core <- c(
  "K02586", # nifH
  "K10944", # amoA
  "K10535", # hao
  "K00368", # nirK
  "K15864", # nirS
  "K04561", # norB
  "K00376", # nosZ
  "K03385", # nrfA
  "K20935"  # hdh
)

ko_n_core <- ko %>%
  mutate(KO = sub("ko:", "", KO_ID)) %>%
  filter(KO %in% nitrogen_kos_core)

# keep only abundance columns and convert to matrix for downstream analysis
ko_n_mat <- ko_n %>%
  dplyr::select(-description, -KO) %>%
  column_to_rownames("KO_ID") %>%
  as.matrix()

ko_n_core_mat <- ko_n_core %>%
  dplyr::select(-description, -KO) %>%
  column_to_rownames("KO_ID") %>%
  as.matrix()

# Normalise abundances
ko_rel <- apply(ko_n_mat, 2, function(x) x / sum(x))
```

### Visualise overall KO differences

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

top_ids <- names(sort(rowMeans(ko_n_mat), decreasing = TRUE))

top_mat <- ko_rel[top_ids, ]

top_labels <- ko_info %>%
  filter(KO_ID %in% top_ids) %>%
  distinct(KO_ID, description)

# make sure label order matches matrix row order
top_labels <- top_labels %>%
  slice(match(rownames(top_mat), KO_ID))

rownames(top_mat) <- paste(top_labels$description)

sample_annot <- metadata %>%
  filter(Name %in% colnames(top_mat)) %>%
  column_to_rownames("Name") %>%
  dplyr::select(Depth, Fjord, Water_mass)

ph <- pheatmap(
  top_mat,
  annotation_col = sample_annot,
  fontsize_col = 8,   # smaller column labels
  angle_col = 45      # rotate labels
)

```


Below is the list of the samples that belong to the cluster on the right side of the heatmap figure above: 

```{r, message=FALSE, warning=FALSE}
cl <- cutree(ph$tree_col, k = 2)
ordered_samples <- colnames(top_mat)[ph$tree_col$order]
right_samples_ko <- ordered_samples[ordered_samples %in% names(cl[cl == 1])]

right_samples_ko
```

Let's see how many of these samples clustered in the right side of this plot are the same as those also cluster at the right of the heatmap of pathways:

```{r, message=FALSE, warning=FALSE}
common_samples <- intersect(right_samples_ko, right_samples_path)
common_samples
```
It seems that all the 20 samples from the right-cluster of the pathway heatmap are again present here. What do the have in common? Likely their taxonomical diversity will be similar. I could group a variable to separate these from the rest of the samples and look at their spatial distribution...

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

top_ids <- names(sort(rowMeans(ko_n_core_mat), decreasing = TRUE))

top_mat <- ko_rel[top_ids, ]

top_labels <- ko_info %>%
  filter(KO_ID %in% top_ids) %>%
  distinct(KO_ID, description)

# make sure label order matches matrix row order
top_labels <- top_labels %>%
  slice(match(rownames(top_mat), KO_ID))

rownames(top_mat) <- paste(top_labels$description)

sample_annot <- metadata %>%
  filter(Name %in% colnames(top_mat)) %>%
  column_to_rownames("Name") %>%
  dplyr::select(Depth, Fjord, Water_mass)

ph <- pheatmap(
  top_mat,
  annotation_col = sample_annot,
  fontsize_col = 8,   # smaller column labels
  angle_col = 45      # rotate labels
)

```

### Using ggpicrust2

ggpicrust2 was built specifically for PICRUSt2-derived functional profiles and provides a clean way to do differential abundance testing, annotation, and visualization of KO / EC / pathway outputs in R.


``` r
# Run ggpicrust2 with input file path
results_file_input <- ggpicrust2(file = abundance_file,
                                 metadata = metadata,
                                 group = "your_group_column", # For example dataset, group = "Environment"
                                 pathway = "KO",
                                 daa_method = "LinDA",
                                 ko_to_kegg = TRUE,
                                 order = "pathway_class",
                                 p_values_bar = TRUE,
                                 x_lab = "pathway_name")

# Run ggpicrust2 with imported data.frame
abundance_data <- read_delim(abundance_file, delim = "\t", col_names = TRUE, trim_ws = TRUE)

# Run ggpicrust2 with input data
results_data_input <- ggpicrust2(data = abundance_data,
                                 metadata = metadata,
                                 group = "your_group_column", # For example dataset, group = "Environment"
                                 pathway = "KO",
                                 daa_method = "LinDA",
                                 ko_to_kegg = TRUE,
                                 order = "pathway_class",
                                 p_values_bar = TRUE,
                                 x_lab = "pathway_name")

# Access the plot and results dataframe for the first DA method
example_plot <- results_file_input[[1]]$plot
example_results <- results_file_input[[1]]$results

# Use the example data in ggpicrust2 package
data(ko_abundance)
data(metadata)
results_file_input <- ggpicrust2(data = ko_abundance,
                                 metadata = metadata,
                                 group = "Environment",
                                 pathway = "KO",
                                 daa_method = "LinDA",
                                 ko_to_kegg = TRUE,
                                 order = "pathway_class",
                                 p_values_bar = TRUE,
                                 x_lab = "pathway_name")

# Analyze the EC or MetaCyc pathway
data(metacyc_abundance)
results_file_input <- ggpicrust2(data = metacyc_abundance,
                                 metadata = metadata,
                                 group = "Environment",
                                 pathway = "MetaCyc",
                                 daa_method = "LinDA",
                                 ko_to_kegg = FALSE,
                                 order = "group",
                                 p_values_bar = TRUE,
                                 x_lab = "description")
results_file_input[[1]]$plot
results_file_input[[1]]$results
```
