# Combined 16S, 18S and metadata

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# ctd <- read_csv("data/CTD_data_Greenland_fjords.csv")
# 
# target_depths <- c(1, 15, 45)
# 
# ctd_selected <- ctd %>%
#   tidyr::crossing(target_depth = target_depths) %>%   # make 3 copies per row
#   group_by(station_idx, station_lab, target_depth) %>% # <-- key fix
#   slice_min(order_by = abs(pressure - target_depth), n = 1, with_ties = FALSE) %>%
#   ungroup()
# 
# ctd_selected %>% head(10)

metadata <- read_csv("data/Metadata_Augmentum.csv")
ps_16s <- readRDS("data/ps_16s.rds")
ps_18s <- readRDS("data/ps_18s.rds")

```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

meta_long <- metadata %>%
  select(10:23) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  )

ggplot(meta_long, aes(y = Value)) +
  geom_boxplot() +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal()

```

::: {.callout-note}
### 💬 Discussion — 


:::

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

ggplot(metadata, aes(x = Salinity, y = Pot_temperature, color = Depth)) +
  geom_point() +
  facet_wrap(~Area) +
  theme_minimal()


```



## Looking for associations

```{r echo=FALSE, message=FALSE, warning=FALSE}

# options(repos = c(
#      bioc = "https://bioc.r-universe.dev",
#      CRAN = "https://cloud.r-project.org"))
# install.packages("SpiecEasi")

library(phyloseq)

# Collapse taxonomic rank
ps16_genus <- tax_glom(ps_16s, taxrank = "Genus")
ps18_genus <- tax_glom(ps_18s, taxrank = "Genus")

# Filter out rare taxa
prev_filter <- function(ps, threshold = 0.2){
  otu <- as(otu_table(ps), "matrix")
  if(!taxa_are_rows(ps)){
    otu <- t(otu)
  }
  
  prevalence <- rowSums(otu > 0) / ncol(otu)
  keep <- prevalence >= threshold
  
  prune_taxa(keep, ps)
}

ps16_filt <- prev_filter(ps16_genus, threshold = 0.2)
ps18_filt <- prev_filter(ps18_genus, threshold = 0.2)

taxa_names(ps16_filt) <- paste0("Bac_", taxa_names(ps16_filt))
taxa_names(ps18_filt) <- paste0("Euk_", taxa_names(ps18_filt))

# Combine matrices

mat16 <- as(otu_table(ps16_filt), "matrix")
mat18 <- as(otu_table(ps18_filt), "matrix")

if(!taxa_are_rows(ps16_filt)) mat16 <- t(mat16)
if(!taxa_are_rows(ps18_filt)) mat18 <- t(mat18)

combined_matrix <- rbind(mat16, mat18)
combined_matrix <- t(combined_matrix)
taxa_names_combined <- colnames(combined_matrix)

library(SpiecEasi)

se <- spiec.easi(
  combined_matrix,
  method = "mb",
  lambda.min.ratio = 1e-2,
  nlambda = 20,
  sel.criterion = "stars",
  pulsar.params = list(rep.num = 50)
)

adj <- getRefit(se)
sum(adj) / 2   # number of edges
rownames(adj) <- taxa_names_combined
colnames(adj) <- taxa_names_combined

```

### Plotting the Network graph (iGraph)

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(igraph)

graph <- graph_from_adjacency_matrix(
  adj,
  mode = "undirected",
  diag = FALSE
)

# Label nodes by domain

V(graph)$domain <- ifelse(
  grepl("^Bac_", V(graph)$name),
  "Bacteria",
  "Eukaryote"
)

table(V(graph)$domain)

# Get cross-domain edges

edges <- as_data_frame(graph, what = "edges")

nodes <- tibble(
  name = V(graph)$name,
  domain = V(graph)$domain
)

edges2 <- edges %>%
  left_join(nodes, by = c("from" = "name")) %>%
  rename(domain_from = domain) %>%
  left_join(nodes, by = c("to" = "name")) %>%
  rename(domain_to = domain)

cross_edges <- edges2 %>% filter(domain_from != domain_to)

nrow(edges2)        # total edges (should match ~725)
nrow(cross_edges)   # cross-domain edges

```

::: {.callout-note}

💬 Note — Cross-domain associations

A total of **278 cross-domain edges** were identified out of **725 total inferred associations**, representing approximately **38%** of all network connections.

This substantial proportion indicates that **bacterial and eukaryotic communities are not structured independently**. Instead, the network reveals meaningful cross-domain covariance patterns, suggesting coordinated ecological dynamics. These associations may reflect bloom-driven bacterial responses, shared environmental filtering, or potential trophic and metabolic interactions between microbial eukaryotes and prokaryotes.
:::

```{r echo=FALSE, message=FALSE, warning=FALSE}
deg <- degree(graph)
cat("🔹 Overall degre of connectivity:\n")
summary(deg)
cat("🔹 Bacteria degre of connectivity:\n")
deg_cross <- degree(graph, v = V(graph)[V(graph)$domain == "Bacteria"])
summary(deg_cross)
cat("🔹 Eukaryote degre of connectivity:\n")
deg_cross <- degree(graph, v = V(graph)[V(graph)$domain == "Eukaryote"])
summary(deg_cross)

```

:: {.callout-note}

💬 Note — Network connectivity structure

The inferred cross-domain network comprised 238 taxa connected by 725 edges, with a median node degree of 6. Bacterial genera exhibited slightly higher connectivity (mean degree ≈ 7, maximum = 19) compared to eukaryotic genera (mean degree ≈ 5.3, maximum = 12). This pattern indicates a moderately connected but sparse network structure, without dominance by a few extreme hubs. The higher connectivity observed among bacteria suggests that prokaryotic taxa may act as integrators within the microbial community, associating with multiple eukaryotic groups. Overall, the degree distribution supports the presence of structured cross-domain coupling rather than random co-occurrence.
:::

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Which taxa are the hubs?
deg_df <- data.frame(
  taxon = V(graph)$name,
  domain = V(graph)$domain,
  degree = degree(graph)
)

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# 1) Extract taxonomy with correct keys (rownames are the taxa_names)
tax16 <- as.data.frame(tax_table(ps16_filt))
tax16$taxon <- rownames(tax16)

tax18 <- as.data.frame(tax_table(ps18_filt))
tax18$taxon <- rownames(tax18)

tax_all <- bind_rows(tax16, tax18)

# 2) Join taxonomy onto your degree table by 'taxon'
deg_tax <- deg_df %>%
  left_join(tax_all, by = "taxon")

# 3) Fallback if Genus is missing
deg_tax <- deg_tax %>%
  mutate(
    Genus_final = case_when(
      !is.na(Genus) & Genus != "" ~ Genus,
      !is.na(Family) & Family != "" ~ paste0("Unclassified_", Family),
      !is.na(Order) & Order != "" ~ paste0("Unclassified_", Order),
      TRUE ~ "Unclassified"
    )
  )

# 4) View top hubs with taxonomy
deg_tax %>%
  arrange(desc(degree)) %>%
  select(taxon, domain, degree, Phylum, Order, Genus_final) %>%
  head(20)

```

### Building Network graph

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(igraph)
library(tidygraph)
library(ggraph)

# Ensure graph has domain attribute
V(graph)$domain <- ifelse(grepl("^Bac_", V(graph)$name), "Bacteria", "Eukaryote")

# Edge list with endpoints
ed <- as_data_frame(graph, what = "edges")

# Domains per endpoint
nodes_df <- tibble(name = V(graph)$name, domain = V(graph)$domain)

ed2 <- ed %>%
  left_join(nodes_df, by = c("from" = "name")) %>% rename(domain_from = domain) %>%
  left_join(nodes_df, by = c("to" = "name")) %>% rename(domain_to   = domain)

# Keep only cross-domain edges
cross_ed <- ed2 %>% filter(domain_from != domain_to)

# Keep only nodes involved in cross-domain edges
cross_nodes <- unique(c(cross_ed$from, cross_ed$to))

graph_cross <- induced_subgraph(graph, vids = cross_nodes)

# Remove isolates just in case
graph_cross <- delete_vertices(graph_cross, degree(graph_cross) == 0)

# Compute degrees in the cross network
deg_cross <- degree(graph_cross)

# Label top hubs (adjust n if needed)
top_n <- 12
top_nodes <- names(sort(deg_cross, decreasing = TRUE))[1:top_n]

V(graph_cross)$label <- ifelse(V(graph_cross)$name %in% top_nodes, V(graph_cross)$name, NA)
V(graph_cross)$size  <- scales::rescale(deg_cross, to = c(2, 8))  # node size by degree

set.seed(1)

ggraph(as_tbl_graph(graph_cross), layout = "fr") +
  geom_edge_link(alpha = 0.25, linewidth = 0.3) +
  geom_node_point(aes(size = size, color = domain), alpha = 0.9) +
  geom_node_text(aes(label = label), repel = TRUE, size = 3) +
  scale_size_identity() +
  theme_void() +
  labs(title = "Cross-domain association network (16S–18S)") +
  theme(legend.position = "bottom")

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create lookup table
genus_lookup <- deg_tax %>%
  select(taxon, Genus_final) %>%
  distinct()

# Convert to dataframe
nodes_df <- data.frame(name = V(graph_cross)$name)

# Join genus info
nodes_df <- nodes_df %>%
  left_join(genus_lookup, by = c("name" = "taxon"))

# Assign genus label to graph
V(graph_cross)$genus <- nodes_df$Genus_final

V(graph_cross)$genus_clean <- gsub("^Unclassified_", "", V(graph_cross)$genus)

deg_cross <- degree(graph_cross)

top_n <- 12
top_nodes <- names(sort(deg_cross, decreasing = TRUE))[1:top_n]

V(graph_cross)$label <- ifelse(
  V(graph_cross)$name %in% top_nodes,
  V(graph_cross)$genus_clean,
  NA
)

library(ggraph)
library(tidygraph)

set.seed(1)

ggraph(as_tbl_graph(graph_cross), layout = "fr") +
  geom_edge_link(alpha = 0.25, linewidth = 0.3) +
  geom_node_point(aes(color = domain, size = deg_cross), alpha = 0.9) +
  geom_node_text(aes(label = label), repel = TRUE, size = 3) +
  scale_size(range = c(2, 8)) +
  theme_void() +
  labs(
    title = "Cross-domain association network (Genus level)",
    color = "Domain"
  ) +
  theme(legend.position = "bottom")

```

### Interactive plot

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(visNetwork)

# Choose which graph to explore:
g <- graph_cross   # or use graph for full network

# Node table
nodes <- data.frame(
  id    = V(g)$name,
  label = ifelse(!is.na(V(g)$genus_clean), V(g)$genus_clean, V(g)$name),
  group = V(g)$domain,
  value = degree(g)  # node size in visNetwork
)

# Edge table
edges <- as_data_frame(g, what = "edges") %>%
  transmute(from = from, to = to)


visNetwork(nodes, edges, height = "800px", width = "100%") %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE
  ) %>%
  visInteraction(
    navigationButtons = TRUE,
    zoomView = TRUE
  ) %>%
  visPhysics(stabilization = TRUE)

```


::: {.callout-note}

::: 
