# Combined 16S, 18S and metadata

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

# ctd <- read_csv("data/CTD_data_Greenland_fjords.csv")
# 
# target_depths <- c(1, 15, 45)
# 
# ctd_selected <- ctd %>%
#   tidyr::crossing(target_depth = target_depths) %>%   # make 3 copies per row
#   group_by(station_idx, station_lab, target_depth) %>% # <-- key fix
#   slice_min(order_by = abs(pressure - target_depth), n = 1, with_ties = FALSE) %>%
#   ungroup()
# 
# ctd_selected %>% head(10)

metadata <- read_csv("data/Metadata_Augmentum.csv")
ps_16s <- readRDS("data/ps_16s.rds")
ps_18s <- readRDS("data/ps_18s.rds")

```

### Combining datasets 

```{r, message=FALSE, warning=FALSE}
library(phyloseq)
library(vegan)

# Collapse taxonomic rank
ps16_genus <- tax_glom(ps_16s, taxrank = "Genus")
ps18_genus <- tax_glom(ps_18s, taxrank = "Genus")

# Filter out rare taxa
prev_filter <- function(ps, threshold = 0.2){
  otu <- as(otu_table(ps), "matrix")
  if(!taxa_are_rows(ps)){
    otu <- t(otu)
  }
  
  prevalence <- rowSums(otu > 0) / ncol(otu)
  keep <- prevalence >= threshold
  
  prune_taxa(keep, ps)
}

ps16_filt <- prev_filter(ps16_genus, threshold = 0.2)
ps18_filt <- prev_filter(ps18_genus, threshold = 0.2)

taxa_names(ps16_filt) <- paste0("Bac_", taxa_names(ps16_filt))
taxa_names(ps18_filt) <- paste0("Euk_", taxa_names(ps18_filt))

# Combine matrices

mat16 <- as(otu_table(ps16_filt), "matrix")
mat18 <- as(otu_table(ps18_filt), "matrix")

combined_matrix <- cbind(mat16, mat18)
taxa_names_combined <- colnames(combined_matrix)

# Create a matching column in metadata

metadata <- metadata %>%
  mutate(
    ps_id = sprintf("%02d_St%d_D%d", row_number(), StationID, Depth)
  )

sd_16s <- sample_data(ps_16s) %>%
  data.frame() %>%
  rownames_to_column("ps_id")

sd_16s_aug <- sd_16s %>%
  left_join(
    metadata %>% select(01:24,52) %>%
      select(-Cruise, -Date, -Area, -StationID, -Depth, -`Water mass`),
    by = "ps_id"
  )

# Put rownames back
sd_16s_aug <- sd_16s_aug %>%
  column_to_rownames("ps_id")

# Apply same sample metadata to both objects
sample_data(ps_16s) <- sample_data(sd_16s_aug)
sample_data(ps_18s) <- sample_data(sd_16s_aug)

# Also update filtered objects
sample_data(ps16_filt) <- sample_data(sd_16s_aug[rownames(sample_data(ps16_filt)), , drop = FALSE])
sample_data(ps18_filt) <- sample_data(sd_16s_aug[rownames(sample_data(ps18_filt)), , drop = FALSE])

# Transform data using Hellinger transformation (best for PCA/RDA):

otu_16s <- as.data.frame(otu_table(ps16_filt))
otu_18s <- as.data.frame(otu_table(ps18_filt))

otu_16s_hel <- decostand(otu_16s, method = "hellinger")
otu_18s_hel <- decostand(otu_18s, method = "hellinger")

# Prepare environmental matrix
env <- metadata %>%
  select(Depth, Salinity, Pot_temperature, NO3, PO4, SiO2, Chla_ugL) %>%
  drop_na()

# scale variables
env_scaled <- scale(env)
```

### Testing Specific questions
#### Is Nitrosopumilus linked to NO₂?

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
nitro_taxa <- taxa_names(ps16_filt)[tax_table(ps16_filt)[, "Genus"] == "Candidatus Nitrosopumilus"]

# OTU table as sample x taxa
otu <- as(otu_table(ps16_filt), "matrix")

if (taxa_are_rows(ps16_filt)) {
  otu <- t(otu)
}

otu <- as.data.frame(otu)

# 3. Extract abundance for that taxon
# rowSums works whether there is one taxon or more than one
nitro_abundance <- rowSums(otu[, nitro_taxa, drop = FALSE])

meta <- sample_data(ps16_filt) %>%
  data.frame() %>%
  rownames_to_column("SampleID")

df <- data.frame(
  SampleID = rownames(otu),
  Nitrosopumilus_count = nitro_abundance
) %>%
  left_join(meta %>% select(SampleID, NO2, Depth, NO3, DOC, Fjord, Pot_temperature), by = "SampleID") %>%
  tidyr::drop_na()

# Relative abundance transform
otu_rel <- decostand(otu, method = "total")

# Extract relative abundance of Nitrosopumilus
nitro_rel <- rowSums(otu_rel[, nitro_taxa, drop = FALSE])

# Build dataframe
df <- data.frame(
  SampleID = rownames(otu_rel),
  Nitrosopumilus_rel = nitro_rel
) %>%
  left_join(meta %>% select(SampleID, NO2, Depth, NO3, DOC, Fjord, Pot_temperature), by = "SampleID") %>%
  tidyr::drop_na()

ggplot(df, aes(x = NO2, y = Nitrosopumilus_rel)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_bw() +
  labs(
    x = "NO2",
    y = "Relative abundance of Candidatus Nitrosopumilus",
    title = "Nitrosopumilus vs NO2"
  )

```

```{r, message=FALSE, warning=FALSE}

cor.test(df$NO2, df$Nitrosopumilus_rel, method = "spearman")

summary(lm(Nitrosopumilus_rel ~ NO2 + Depth, data = df))
```
```{r, message=FALSE, warning=FALSE}
ggplot(df, aes(x = NO2, y = Nitrosopumilus_rel, color = Depth)) +
     geom_point(size = 3) +
     geom_smooth(method = "lm", se = FALSE) +
     theme_bw()
```
The relationship is depth-dependent. No relationship at depth 45!

```{r, message=FALSE, warning=FALSE}
summary(lm(Nitrosopumilus_rel ~ NO2 * Depth, data = df))
```

Candidatus Nitrosopumilus abundance is positively associated with NO₂ concentrations, but this relationship is depth-dependent. The correlation is most pronounced at intermediate depth (15 m), suggesting a zone of active ammonia-oxidation, while deeper waters show high Nitrosopumilus abundance independent of NO₂, possibly reflecting steady-state nitrogen cycling (coupling of nitrification reactions) or different environmental controls.

#### Using counts instead of relative abundance (better approach)

```{r, message=FALSE, warning=FALSE}
# OTU table as sample x taxa
otu <- as(otu_table(ps16_filt), "matrix")
if (taxa_are_rows(ps16_filt)) {
  otu <- t(otu)
}
otu <- as.data.frame(otu)

# Nitrosopumilus taxa
nitro_taxa <- taxa_names(ps16_filt)[tax_table(ps16_filt)[, "Genus"] == "Candidatus Nitrosopumilus"]

# Counts
nitro_abundance <- rowSums(otu[, nitro_taxa, drop = FALSE])

# Total reads per sample
total_reads <- rowSums(otu)

# Metadata
meta <- sample_data(ps16_filt) %>%
  data.frame() %>%
  rownames_to_column("SampleID")

# Build clean dataframe
df_counts <- data.frame(
  SampleID = rownames(otu),
  Nitrosopumilus_count = nitro_abundance,
  total_reads = total_reads
) %>%
  left_join(meta %>% select(SampleID, NO2, Depth, NO3, DOC, Fjord, Pot_temperature), by = "SampleID") %>%
  tidyr::drop_na()

head(df_counts)
nrow(df_counts)

library(MASS)

m_nb <- glm.nb(
  Nitrosopumilus_count ~ NO2 * Depth + offset(log(total_reads)),
  data = df_counts
)

summary(m_nb)
```

Using counts with a negative binomial model (and a library-size offset) gives you a more faithful representation of abundance than relative proportions. The negative binomial model didn’t change our conclusions — it strengthened them and made them more defensible.


#### Do we have nitrite oxidizing bacteria?

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
nitrospina_taxa <- "Bac_ASV483"

nitrospina_rel <- otu_rel[, nitrospina_taxa]

df_combined <- df %>%
  left_join(
    data.frame(
      SampleID = rownames(otu_rel),
      Nitrospina_rel = nitrospina_rel
    ),
    by = "SampleID"
  )

ggplot(df_combined, aes(x = NO2, y = Nitrospina_rel, color = Depth)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(
    y = "Nitrospinaceae (LS-NOB)",
    title = "NOB vs NO2"
  )

```

```{r, message=FALSE, warning=FALSE}
ggplot(df_combined, aes(x = Nitrosopumilus_rel, y = Nitrospina_rel, color = Depth)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(
    x = "Nitrosopumilus (AOA)",
    y = "Nitrospinaceae (NOB)",
    title = "Coupling between ammonia and nitrite oxidizers"
  )
```
This plot suggest strong AOA–NOB coupling! Their co-variation means that they are functioning as a linked nitrification system. Let's test it statistically:

```{r, message=FALSE, warning=FALSE}
cor.test(df_combined$Nitrosopumilus_rel,
         df_combined$Nitrospina_rel,
         method = "spearman")
```
This result indicates that Nitrosopumilus and Nitrospinaceae (NOB) taxa co-vary strongly across samples.

#### Is NO3 a predictor of Nitrospinaceae (NOB) abundance?

```{r, message=FALSE, warning=FALSE}

cor.test(df_combined$NO3, df_combined$Nitrospina_rel, method = "spearman")

summary(lm(Nitrospina_rel ~ NO3 + Depth, data = df_combined))
```
```{r, message=FALSE, warning=FALSE}
ggplot(df_combined, aes(x = NO3, y = Nitrospina_rel, color = Depth)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(
    y = "Nitrospinaceae (LS-NOB)",
    title = "NOB vs NO3"
  )
```

Nitrospinaceae abundance is strongly associated with NO₃ concentrations, consistent with its role as a nitrite oxidizer. This relationship is strongest at depth, suggesting that nitrite oxidation and nitrate accumulation occur primarily in deeper waters. In contrast, intermediate depths show stronger coupling between ammonia oxidizers and NO₂, indicating spatial separation between nitrification steps.

### RDA

```{r, message=FALSE, warning=FALSE}
rda_16s <- rda(otu_16s_hel ~ Depth + Salinity + Pot_temperature + NO3 + PO4 + SiO2 + Chla_ugL, data = metadata)
rda_18s <- rda(otu_18s_hel ~ Depth + Salinity + Pot_temperature + NO3 + PO4 + SiO2 + Chla_ugL, data = metadata)

# Test significance

anova(rda_16s, by = "term")

anova(rda_18s, by = "term")

```

```{r, message=FALSE, warning=FALSE}
# Extract scores

scores_16s <- scores(rda_16s, display = "sites") %>% as.data.frame()
scores_16s <- cbind(scores_16s, metadata)

env_arrows <- scores(rda_16s, display = "bp") %>% as.data.frame()

eig <- rda_16s$CCA$eig
pct <- round(100 * eig / sum(eig), 1)

x_lab <- paste0("RDA1 (", pct[1], "%)")
y_lab <- paste0("RDA2 (", pct[2], "%)")

scores_16s$Depth <- as.factor(scores_16s$Depth)

ggplot(scores_16s, aes(RDA1, RDA2)) +
  geom_point(aes(color = Depth, shape = Area), size = 3) + 
  ggrepel::geom_text_repel(aes(label = StationID), size = 3) +
  geom_segment(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = RDA1, y = RDA2, label = rownames(env_arrows)),
    size = 4
  ) +
  theme_minimal() + ggtitle("RDA of 16S Bacterial Communities Along Environmental Gradients") +
  labs(
    x = x_lab,
    y = y_lab
  )

```

```{r, message=FALSE, warning=FALSE}
# Extract scores

scores_18s <- scores(rda_18s, display = "sites") %>% as.data.frame()
scores_18s <- cbind(scores_18s, metadata)

env_arrows <- scores(rda_18s, display = "bp") %>% as.data.frame()

eig <- rda_18s$CCA$eig
pct <- round(100 * eig / sum(eig), 1)

x_lab <- paste0("RDA1 (", pct[1], "%)")
y_lab <- paste0("RDA2 (", pct[2], "%)")

scores_18s$Depth <- as.factor(scores_18s$Depth)

ggplot(scores_18s, aes(RDA1, RDA2)) +
  geom_point(aes(color = Depth, shape = Area), size = 3) + 
  ggrepel::geom_text_repel(aes(label = StationID), size = 3) +
  geom_segment(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = RDA1, y = RDA2, label = rownames(env_arrows)),
    size = 4
  ) +
  theme_minimal() + ggtitle("RDA of 18S eukaryotic Communities Along Environmental Gradients")+
  labs(
    x = x_lab,
    y = y_lab
  )

```

**Note**: What is collinearity?
definition: Two or more environmental variables are strongly correlated with each other.

As part of a collinearity diagnostics, we can investigate the correlations between variables and the Variance Inflation Factor (VIF).

Let's start with the correlation:

```{r, message=FALSE, warning=FALSE}
env <- metadata %>%
  dplyr::select(Depth, Pot_temperature, Salinity, NO3,NO2, SiO2, PO4, DOC, Chla_ugL, a350, PARAFAC1,PARAFAC2, PARAFAC3) %>% tidyr::drop_na()

library(corrplot)

cor_mat <- cor(env, use = "complete.obs")
corrplot(cor_mat, method = "color")
```

We can test VIF with vif.cca():

:::{.column width="50%"}

| VIF value | Meaning                  |
| --------- | ------------------------ |
| < 5       | OK                       |
| 5–10      | Moderate collinearity ⚠️ |
| > 10      | Strong collinearity ❌    |
:::

#### Calculating the VIF of the variables used:
##### 16S

:::{.column width="50%"}
```{r, message=FALSE, warning=FALSE}
library(knitr)

vif_values <- vif.cca(rda_16s)

kable(
  data.frame(
    Variable = names(vif_values),
    VIF = as.numeric(vif_values),
    row.names = NULL
  ),
  digits = 2,
  caption = "Variance Inflation Factors (VIF) for predictor variables"
)
```
:::


:::{.column width="50%"}
| Variable         | VIF  | Interpretation     |
|------------------|------|--------------------|
| Depth            | 3.7  | ✅ OK              |
| Salinity         | 18.5 | ❌ Very high       |
| Pot_temperature  | 2.3  | ✅ OK              |
| NO3              | 5.9  | ⚠️ Borderline      |
| PO4              | 13.8 | ❌ High            |
| SiO2             | 8.3  | ⚠️ Moderate        |
| Chla             | 2.0  | ✅ OK              |
:::

What to do? We could remove redundant variables

For example: keeping these:

- Depth → physical structure
- Temperature → water mass
- NO3 → nutrients
- Chla → biology

```{r, message=FALSE, warning=FALSE}
rda_16s_clean <- rda(
  otu_16s_hel ~ Depth + Pot_temperature + NO2 + PARAFAC1 + PARAFAC3,
  data = metadata, na.action = na.omit
)

anova(rda_16s_clean, by = "term")
```

:::{.column width="50%"}
```{r, message=FALSE, warning=FALSE}
vif_values <- vif.cca(rda_16s_clean)

kable(
  data.frame(
    Variable = names(vif_values),
    VIF = as.numeric(vif_values),
    row.names = NULL
  ),
  digits = 2,
  caption = "Variance Inflation Factors (VIF) for predictor variables"
)
```
:::

```{r, message=FALSE, warning=FALSE}
# Extract scores

scores_16s <- scores(rda_16s_clean, display = "sites") %>% as.data.frame()
meta_used <- metadata 
# %>%
#   drop_na(DOC)
scores_16s <- cbind(scores_16s, meta_used)

env_arrows <- scores(rda_16s_clean, display = "bp") %>% as.data.frame()

eig <- rda_16s_clean$CCA$eig
pct <- round(100 * eig / sum(eig), 1)

x_lab <- paste0("RDA1 (", pct[1], "%)")
y_lab <- paste0("RDA2 (", pct[2], "%)")

scores_16s$Depth <- as.factor(scores_16s$Depth)

ggplot(scores_16s, aes(RDA1, RDA2)) +
  geom_point(aes(color = Depth, shape = Area), size = 3) + 
  ggrepel::geom_text_repel(aes(label = StationID), size = 3) +
  geom_segment(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = RDA1, y = RDA2, label = rownames(env_arrows)),
    size = 4
  ) +
  theme_minimal() + ggtitle("RDA of 16S Bacterial Communities Along Environmental Gradients")+
  labs(
    x = x_lab,
    y = y_lab
  )

```

##### 18S

```{r, message=FALSE, warning=FALSE}
rda_18s_clean <- rda(
  otu_18s_hel ~ Depth + Pot_temperature + NO2 + PARAFAC1 + PARAFAC3,
  data = metadata, na.action = na.omit
)

anova(rda_18s_clean, by = "term")
```

:::{.column width="50%"}
```{r, message=FALSE, warning=FALSE}
vif_values <- vif.cca(rda_18s_clean)

kable(
  data.frame(
    Variable = names(vif_values),
    VIF = as.numeric(vif_values),
    row.names = NULL
  ),
  digits = 2,
  caption = "Variance Inflation Factors (VIF) for predictor variables"
)
```
:::

```{r, message=FALSE, warning=FALSE}
# Extract scores

scores_18s <- scores(rda_18s_clean, display = "sites") %>% as.data.frame()
meta_used <- metadata 
# %>%
#   drop_na(DOC)
scores_18s <- cbind(scores_18s, meta_used)

env_arrows <- scores(rda_18s_clean, display = "bp") %>% as.data.frame()

eig <- rda_18s_clean$CCA$eig
pct <- round(100 * eig / sum(eig), 1)

x_lab <- paste0("RDA1 (", pct[1], "%)")
y_lab <- paste0("RDA2 (", pct[2], "%)")

scores_18s$Depth <- as.factor(scores_18s$Depth)

ggplot(scores_18s, aes(RDA1, RDA2)) +
  geom_point(aes(color = Depth, shape = Area), size = 3) + 
  ggrepel::geom_text_repel(aes(label = StationID), size = 3) +
  geom_segment(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "black"
  ) +
  geom_text(
    data = env_arrows,
    inherit.aes = FALSE,
    aes(x = RDA1, y = RDA2, label = rownames(env_arrows)),
    size = 4
  ) +
  theme_minimal() + ggtitle("RDA of 18S planktonic eukaryotic Communities Along Environmental Gradients")+
  labs(
    x = x_lab,
    y = y_lab
  )
```


### Which taxa covariate with environmental gradients in our RDAs?


1. Extract and Plot Loadings from dbRDA

```{r, message=FALSE, warning=FALSE}
taxa_scores <- scores(rda_16s_clean, display = "species",  scaling = 2) %>%
  as.data.frame() %>%
  rownames_to_column("Genus")


env_arrows <- scores(rda_16s_clean, display = "bp", scaling = 2) %>%
  as.data.frame() %>%
  rownames_to_column("variable")

taxa_scores <- taxa_scores %>%
  mutate(strength = sqrt(RDA1^2 + RDA2^2))

top_taxa <- taxa_scores %>%
  arrange(desc(strength)) %>%
  slice(1:20)

ggplot() +
  # taxa points
  geom_point(data = top_taxa, aes(RDA1, RDA2), color = "blue") +
  geom_text(data = top_taxa, aes(RDA1, RDA2, label = Genus), size = 3) +
  
  # environmental arrows
  geom_segment(data = env_arrows,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "darkgreen") +
  
  geom_text(data = env_arrows,
            aes(x = RDA1, y = RDA2, label = variable),
            color = "darkgreen",
            fontface = "bold") +
  
  theme_minimal() +
  labs(title = "Environmental Drivers of Bacterial Genera (RDA)")

```



2. Correlation Analysis (Taxa vs. Environment)
3. Differential Abundance Testing
4. Machine Learning (Random Forest)

## Looking for associations

```{r, message=FALSE, warning=FALSE}

# options(repos = c(
#      bioc = "https://bioc.r-universe.dev",
#      CRAN = "https://cloud.r-project.org"))
# install.packages("SpiecEasi")

library(SpiecEasi)

se <- spiec.easi(
  combined_matrix,
  method = "mb",
  lambda.min.ratio = 1e-2,
  nlambda = 20,
  sel.criterion = "stars",
  pulsar.params = list(rep.num = 50)
)

cat("Number of edges:\n")

adj <- getRefit(se)
sum(adj) / 2   # number of edges
rownames(adj) <- taxa_names_combined
colnames(adj) <- taxa_names_combined

```

### Plotting the Network graph (iGraph)

```{r, message=FALSE, warning=FALSE}

library(igraph)

graph <- graph_from_adjacency_matrix(
  adj,
  mode = "undirected",
  diag = FALSE
)

# Label nodes by domain

V(graph)$domain <- ifelse(
  grepl("^Bac_", V(graph)$name),
  "Bacteria",
  "Eukaryote"
)

table(V(graph)$domain)

# Get cross-domain edges

edges <- as_data_frame(graph, what = "edges")

nodes <- tibble(
  name = V(graph)$name,
  domain = V(graph)$domain
)

edges2 <- edges %>%
  left_join(nodes, by = c("from" = "name")) %>%
  rename(domain_from = domain) %>%
  left_join(nodes, by = c("to" = "name")) %>%
  rename(domain_to = domain)

cross_edges <- edges2 %>% filter(domain_from != domain_to)

nrow(edges2)        # total edges (should match ~725)
nrow(cross_edges)   # cross-domain edges

```

::: {.callout-note}
💬 Note — Cross-domain associations

A total of **278 cross-domain edges** were identified out of **725 total inferred associations**, representing approximately **38%** of all network connections.

This substantial proportion indicates that **bacterial and eukaryotic communities are not structured independently**. Instead, the network reveals meaningful cross-domain covariance patterns, suggesting coordinated ecological dynamics. These associations may reflect bloom-driven bacterial responses, shared environmental filtering, or potential trophic and metabolic interactions between microbial eukaryotes and prokaryotes.
:::

```{r, message=FALSE, warning=FALSE}
deg <- degree(graph)
cat("🔹 Overall degre of connectivity:\n")
summary(deg)
cat("🔹 Bacteria degre of connectivity:\n")
deg_cross <- degree(graph, v = V(graph)[V(graph)$domain == "Bacteria"])
summary(deg_cross)
cat("🔹 Eukaryote degre of connectivity:\n")
deg_cross <- degree(graph, v = V(graph)[V(graph)$domain == "Eukaryote"])
summary(deg_cross)

```

::: {.callout-note title="Network connectivity structure"}
The inferred cross-domain network comprised 238 taxa connected by 725 edges, with a median node degree of 6. Bacterial genera exhibited slightly higher connectivity (mean degree ≈ 7, maximum = 19) compared to eukaryotic genera (mean degree ≈ 5.3, maximum = 12). This pattern indicates a moderately connected but sparse network structure, without dominance by a few extreme hubs. The higher connectivity observed among bacteria suggests that prokaryotic taxa may act as integrators within the microbial community, associating with multiple eukaryotic groups. Overall, the degree distribution supports the presence of structured cross-domain coupling rather than random co-occurrence.
:::

```{r, message=FALSE, warning=FALSE}
# Which taxa are the hubs?
deg_df <- data.frame(
  taxon = V(graph)$name,
  domain = V(graph)$domain,
  degree = degree(graph)
)

```

:::{.column width="100%"}
```{r, message=FALSE, warning=FALSE}

# 1) Extract taxonomy with correct keys (rownames are the taxa_names)
tax16 <- as.data.frame(tax_table(ps16_filt))
tax16$taxon <- rownames(tax16)

tax18 <- as.data.frame(tax_table(ps18_filt))
tax18$taxon <- rownames(tax18)

tax_all <- bind_rows(tax16, tax18)

# 2) Join taxonomy onto your degree table by 'taxon'
deg_tax <- deg_df %>%
  left_join(tax_all, by = "taxon")

# 3) Fallback if Genus is missing
deg_tax <- deg_tax %>%
  mutate(
    Genus_final = case_when(
      !is.na(Genus) & Genus != "" ~ Genus,
      !is.na(Family) & Family != "" ~ paste0("Unclassified_", Family),
      !is.na(Order) & Order != "" ~ paste0("Unclassified_", Order),
      TRUE ~ "Unclassified"
    )
  )

# 4) View top hubs with taxonomy
deg_tax %>%
  arrange(desc(degree)) %>%
  dplyr::select(taxon, domain, degree, Phylum, Order, Genus_final) %>%
  head(20) %>% kable() %>% kable_styling(full_width = FALSE, position = "center")

```
:::

### Building Network graph

```{r, message=FALSE, warning=FALSE}
library(igraph)
library(tidygraph)
library(ggraph)

# Ensure graph has domain attribute
V(graph)$domain <- ifelse(grepl("^Bac_", V(graph)$name), "Bacteria", "Eukaryote")

# Edge list with endpoints
ed <- as_data_frame(graph, what = "edges")

# Domains per endpoint
nodes_df <- tibble(name = V(graph)$name, domain = V(graph)$domain)

ed2 <- ed %>%
  left_join(nodes_df, by = c("from" = "name")) %>% rename(domain_from = domain) %>%
  left_join(nodes_df, by = c("to" = "name")) %>% rename(domain_to   = domain)

# Keep only cross-domain edges
cross_ed <- ed2 %>% filter(domain_from != domain_to)

# Keep only nodes involved in cross-domain edges
cross_nodes <- unique(c(cross_ed$from, cross_ed$to))

graph_cross <- induced_subgraph(graph, vids = cross_nodes)

# Remove isolates just in case
graph_cross <- delete_vertices(graph_cross, degree(graph_cross) == 0)

# Compute degrees in the cross network
deg_cross <- degree(graph_cross)

# Create lookup table
genus_lookup <- deg_tax %>%
  dplyr::select(taxon, Genus_final) %>%
  distinct()

# Convert to dataframe
nodes_df <- data.frame(name = V(graph_cross)$name)

# Join genus info
nodes_df <- nodes_df %>%
  left_join(genus_lookup, by = c("name" = "taxon"))

# Assign genus label to graph
V(graph_cross)$genus <- nodes_df$Genus_final

V(graph_cross)$genus_clean <- gsub("^Unclassified_", "", V(graph_cross)$genus)

deg_cross <- degree(graph_cross)

top_n <- 12
top_nodes <- names(sort(deg_cross, decreasing = TRUE))[1:top_n]

V(graph_cross)$label <- ifelse(
  V(graph_cross)$name %in% top_nodes,
  V(graph_cross)$genus_clean,
  NA
)

set.seed(1)

ggraph(as_tbl_graph(graph_cross), layout = "fr") +
  geom_edge_link(alpha = 0.25, linewidth = 0.3) +
  geom_node_point(aes(color = domain, size = deg_cross), alpha = 0.9) +
  geom_node_text(aes(label = label), repel = TRUE, size = 3) +
  scale_size(range = c(2, 8)) +
  theme_void() +
  labs(
    title = "Cross-domain association network (Genus level)",
    color = "Domain"
  ) +
  theme(legend.position = "bottom")

```

### Interactive plots
#### Overall cross-domain network

```{r, message=FALSE, warning=FALSE}

library(visNetwork)

# Choose which graph to explore:
g <- graph_cross   # or use graph for full network

# Node table
nodes <- data.frame(
  id    = V(g)$name,
  label = ifelse(!is.na(V(g)$genus_clean), V(g)$genus_clean, V(g)$name),
  group = V(g)$domain,
  value = degree(g)  # node size in visNetwork
)

# Edge table
edges <- as_data_frame(g, what = "edges") %>%
  transmute(from = from, to = to)


visNetwork(nodes, edges, height = "800px", width = "100%") %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE
  ) %>%
  visInteraction(
    navigationButtons = TRUE,
    zoomView = TRUE
  ) %>%
  visPhysics(stabilization = TRUE)

```


::: {.callout-note}
This is great but it's too complex so let's filter out nodes that have degree < 11.
::: 

#### core of the network

```{r, message=FALSE, warning=FALSE}
# Calculate node degree
deg <- degree(graph_cross)

# Keep nodes with degree > 3
nodes_keep <- names(deg[deg > 10])

# Create filtered graph
graph_filtered <- induced_subgraph(graph_cross, vids = nodes_keep)

# Check new size
vcount(graph_filtered)
ecount(graph_filtered)

# Calculate node degree
deg <- degree(graph_cross)

# Keep nodes with degree > 3
nodes_keep <- names(deg[deg > 10])

# Create filtered graph
graph_filtered <- induced_subgraph(graph_cross, vids = nodes_keep)

# Check new size
vcount(graph_filtered)
ecount(graph_filtered)

nodes <- data.frame(
  id    = V(graph_filtered)$name,
  label = ifelse(!is.na(V(graph_filtered)$genus_clean),
                 V(graph_filtered)$genus_clean,
                 V(graph_filtered)$name),
  group = V(graph_filtered)$domain,
  value = degree(graph_filtered)
)

edges <- as_data_frame(graph_filtered, what = "edges") %>%
  transmute(from = from, to = to)

visNetwork(nodes, edges, height = "800px", width = "100%") %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE
  ) %>%
  visInteraction(
    navigationButtons = TRUE,
    zoomView = TRUE
  ) %>%
  visPhysics(stabilization = TRUE)
```

::: {.callout-note}
Bacterial hubs (blue) dominate the network core. These taxa are typical marine heterotrophic bacteria associated with phytoplankton or organic matter degradation. 

The central structure appears around:
RS62 marine group → Aureispira → Mariniblastus → OM27

Eukaryotes (yellow) appear as satellites linked to bacterial hubs. They connect mainly through bacterial hubs rather than directly to each other.

That pattern suggests that bacterial communities may be responding to eukaryotic blooms or shared environmental drivers.
::: 

#### Cross-domain partner preference

To better understand the ecological structure of the inferred 16S–18S association network, we examined **cross-domain partner preference**, i.e. which bacterial taxa were most frequently associated with eukaryotic taxa and vice versa. This allows identification of potential bacterial hubs linked to particular protistan or metazoan groups, and helps reveal whether the network is structured around specific cross-domain partnerships rather than diffuse associations.

```{r, message=FALSE, warning=FALSE}

# Prepare node metadata from network + taxonomy
nodes_df <- data.frame(
  taxon = V(graph)$name,
  domain = V(graph)$domain,
  degree = degree(graph),
  stringsAsFactors = FALSE
) %>%
  left_join(
    deg_tax %>%
      dplyr::select(taxon, Phylum, Class, Order, Family, Genus_final),
    by = "taxon"
  )

# Clean label column
nodes_df <- nodes_df %>%
  mutate(
    Label = ifelse(!is.na(Genus_final) & Genus_final != "",
                   Genus_final,
                   taxon)
  )

# Prepare edge table and retain cross-domain edges only

edges_df <- as_data_frame(graph, what = "edges") %>%
  left_join(nodes_df %>% dplyr::select(taxon, domain, Label, Phylum, Class, Order, Family),
            by = c("from" = "taxon")) %>%
  rename(domain_from = domain,
         label_from  = Label,
         phylum_from = Phylum,
         class_from  = Class,
         order_from  = Order,
         family_from = Family) %>%
  left_join(nodes_df %>% dplyr::select(taxon, domain, Label, Phylum, Class, Order, Family),
            by = c("to" = "taxon")) %>%
  rename(domain_to = domain,
         label_to  = Label,
         phylum_to = Phylum,
         class_to  = Class,
         order_to  = Order,
         family_to = Family)

cross_edges <- edges_df %>%
  filter(domain_from != domain_to)

nrow(cross_edges)

# Standardize edge direction:
# bacteria always on the left, eukaryotes on the right

cross_pref <- cross_edges %>%
  mutate(
    bacteria = ifelse(domain_from == "Bacteria", label_from, label_to),
    bacteria_family = ifelse(domain_from == "Bacteria", family_from, family_to),
    bacteria_order  = ifelse(domain_from == "Bacteria", order_from, order_to),
    eukaryote = ifelse(domain_from == "Eukaryote", label_from, label_to),
    euk_family = ifelse(domain_from == "Eukaryote", family_from, family_to),
    euk_order  = ifelse(domain_from == "Eukaryote", order_from, order_to)
  )

# Count how many times each bacterial genus is connected
# to each eukaryotic order

partner_pref <- cross_pref %>%
  count(bacteria, euk_order, sort = TRUE) %>%
  filter(!is.na(euk_order))

head(partner_pref, 20)

# Top 20 most frequent bacteria–eukaryote order associations
partner_pref %>%
  arrange(desc(n)) %>%
  slice(1:20)
```
```{r, message=FALSE, warning=FALSE}
# Identify important bacteria
top_bacteria <- partner_pref %>%
  group_by(bacteria) %>%
  summarise(total = sum(n)) %>%
  filter(total >= 2) %>%
  pull(bacteria)

# Identify important eukaryotic orders
top_euk <- partner_pref %>%
  group_by(euk_order) %>%
  summarise(total = sum(n)) %>%
  filter(total >= 2) %>%
  pull(euk_order)

heat_df2 <- partner_pref %>%
  filter(
    bacteria %in% top_bacteria,
    euk_order %in% top_euk
  )

ggplot(heat_df2, aes(x = euk_order, y = bacteria, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Fjord-specific core subnetworks

To explore whether the cross-domain association network is organized differently among fjords, we can visualise **fjord-specific core subnetworks** derived from the global 16S–18S cross-domain network. For each fjord, let's retain taxa present in samples from that fjord, extract the corresponding induced subgraph, and then filter to keep only highly connected nodes (core taxa). This approach provides an exploratory view of the dominant cross-domain network structure within each fjord while avoiding the instability associated with re-inferring sparse networks from small fjord-specific sample sets.

```{r, message=FALSE, warning=FALSE}
# Make sure sample order is identical between filtered 16S and 18S objects
stopifnot(all(sample_names(ps16_filt) == sample_names(ps18_filt)))

# Extract count matrices with taxa in rows
mat16 <- as(otu_table(ps16_filt), "matrix")
mat18 <- as(otu_table(ps18_filt), "matrix")

if (!taxa_are_rows(ps16_filt)) mat16 <- t(mat16)
if (!taxa_are_rows(ps18_filt)) mat18 <- t(mat18)

# Metadata
meta <- data.frame(sample_data(ps16_filt))
meta$Sample <- rownames(meta)

# Fjord names
fjords <- sort(unique(meta$Fjord))
fjords
```
```{r, message=FALSE, warning=FALSE}
make_fjord_network <- function(fjord_name,
                               graph_obj = NULL,
                               meta_df = NULL,
                               mat16_in = NULL,
                               mat18_in = NULL,
                               degree_threshold = 5,
                               min_taxon_sum = 1) {
  
  if (is.null(graph_obj)) graph_obj <- graph_cross
  if (is.null(meta_df)) meta_df <- meta
  if (is.null(mat16_in)) mat16_in <- mat16
  if (is.null(mat18_in)) mat18_in <- mat18
  
  # Samples from that fjord
  samp <- meta_df$Sample[meta_df$Fjord == fjord_name]
  
  # Taxa present in that fjord
  bac_present <- rownames(mat16_in)[rowSums(mat16_in[, samp, drop = FALSE]) >= min_taxon_sum]
  euk_present <- rownames(mat18_in)[rowSums(mat18_in[, samp, drop = FALSE]) >= min_taxon_sum]
  taxa_present <- unique(c(bac_present, euk_present))
  
  # Keep only taxa that are also in the global cross-domain graph
  taxa_present <- intersect(taxa_present, V(graph_obj)$name)
  
  # Induced fjord subgraph
  g_fjord <- induced_subgraph(graph_obj, vids = taxa_present)
  
  # Remove isolated nodes
  g_fjord <- delete_vertices(g_fjord, degree(g_fjord) == 0)
  
  if (vcount(g_fjord) == 0) return(NULL)
  
  # Core filtering by degree within fjord subgraph
  deg_fjord <- degree(g_fjord)
  keep_nodes <- names(deg_fjord[deg_fjord > degree_threshold])
  
  g_core <- induced_subgraph(g_fjord, vids = keep_nodes)
  g_core <- delete_vertices(g_core, degree(g_core) == 0)
  
  if (vcount(g_core) == 0) return(NULL)
  
  # Build nodes table
  nodes <- data.frame(
    id    = V(g_core)$name,
    label = ifelse(!is.na(V(g_core)$genus_clean),
                   V(g_core)$genus_clean,
                   V(g_core)$name),
    group = V(g_core)$domain,
    value = degree(g_core),
    stringsAsFactors = FALSE
  )
  
  nodes$color <- ifelse(nodes$group == "Bacteria", "#2C7FB8", "#D95F0E")
  
  # Build edge table
  edges <- as_data_frame(g_core, what = "edges") %>%
    transmute(from = from, to = to)
  
  visNetwork(nodes, edges, height = "750px", width = "100%") %>%
    visNodes(borderWidth = 1) %>%
    visOptions(
      highlightNearest = TRUE,
      nodesIdSelection = TRUE
    ) %>%
    visInteraction(
      navigationButtons = TRUE,
      zoomView = TRUE,
      dragNodes = TRUE,
      dragView = TRUE
    ) %>%
    visPhysics(stabilization = TRUE) %>%
    visLayout(randomSeed = 7) %>%
    visLegend()
}
```

##### Sermilik
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
make_fjord_network("Sermilik", degree_threshold = 10)
```

##### Iterdalk
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
make_fjord_network("Iterdalk", degree_threshold = 10)
```

##### Tunulliarfik
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
make_fjord_network("Tunulliarfik", degree_threshold = 10)
```